---
title: "CPOX_Manuscript"
date: "2026-4-7"
editor_options: 
  chunk_output_type: inline
---

# =============================================================================
# CPOX Deep Mutational Scanning (DMS) Analysis — Manuscript Code
#
# This R Markdown document contains all analyses for the CPOX DMS manuscript.
# CPOX (coproporphyrinogen oxidase) is the enzyme deficient in hereditary
# coproporphyria (HCP). The experiment measures variant functional scores
# under two conditions:
#   • Baseline  — standard growth selection
#   • Metal     — growth selection in the presence of mercury (Hg)
#
# Chunks covered:
#   1.  Score Distributions
#   2.  Consensus ring map (circular visualisation)
#   3.  Population-level assessment (UK Biobank / gnomAD)
#   4.  Baseline vs. Metal map correlation
#   5.  Delta score analysis (baseline – metal)
#   6.  Functional scores vs. predicted ΔΔG (thermodynamic stability)
#   7.  Correlation with patient enzymatic activity
#   8.  Hypercomplementing variants vs. AlphaMissense
#   9.  Dunham 33-map substitution trend comparison
#   10. Amino acid substitution type analysis (FROM / TO)
#   11. Structure-based analysis (SASA / quaternary domains)
#   12. MTS-region structure analysis
#   13. Negative-charge substitution at second decarboxylation residues
#   14. Delta score impact classification (APR / Improved / Known)
#   15. Delta significance heatmap
#   16. Balanced Precision-Recall Curves (PRC)
#   17. PRC with random predictor
#   18. LLRp evidence for VUS reclassification
#   19. VUS reclassification summary
#   20. Variant validation (growth curves)
# =============================================================================


```{r, warning=FALSE, Libraries}
# Load all libraries used across the manuscript analyses
library(cowplot)
library(data.table)
library(ggnewscale)    
library(ggplot2)
library(ggpubr)
library(ggrepel)
library(ggsignif)
library(ggnewscale)
library(gridExtra)
library(httr)          
library(jsonlite)     
library(mgsub)         
library(patchwork)
library(plotly)
library(purrr)
library(RcppRoll)
library(rstatix)
library(scales)
library(stringr)
library(tidyverse)
library(RColorBrewer)
library(reshape2)
library(yogiroc)       
library(readxl)       
library(zoo)        

```


```{r, warning=FALSE, Data Loading Helper}
# =============================================================================
# HELPER — read_sheet()
#
# Reads a named sheet from Manuscript_Data.xlsx, skipping the first row
# (generic info row), then coerces every column to numeric where possible.
# Character columns (e.g. hgvs_pro, disease_type) are left as-is.
# Use this instead of readxl::read_excel() directly throughout the script.
# =============================================================================

read_sheet <- function(sheet) {
  df <- readxl::read_excel(
    "~/Desktop/Manuscript_Data.xlsx",
    sheet    = sheet,
    skip     = 1,
    na       = c("", "NA", "N/A", "#NAME?", "#VALUE!", "#REF!",
                 "#DIV/0!", "#NULL!", "#NUM!", "#N/A")
  )
  df[] <- lapply(df, function(col) {
    if (is.character(col)) {
      # Map Excel Inf strings to R Inf, then coerce
      col <- trimws(col)
      col[col %in% c("Inf", "inf", "INF")]   <- "Inf"
      col[col %in% c("-Inf", "-inf", "-INF")] <- "-Inf"
      converted <- suppressWarnings(as.numeric(col))
      # Keep as character only if conversion turned non-NA values into NA
      non_na_orig      <- !is.na(col)
      non_na_converted <- !is.na(converted)
      if (all(non_na_converted[non_na_orig])) converted else col
    } else {
      col
    }
  })
  as.data.frame(df)
}
```


```{r, warning=FALSE, Score Distributions}
# =============================================================================
# CHUNK 1 — Score Distributions
#
# Produces mirrored bar-density histograms for each map (Baseline, Metal).
# Missense density is plotted below zero; nonsense and synonymous above.
# Also reports:
#   • Median score per variant class
#   • Percentage of missense variants with score > 1.1 (hyperactive candidates)
# =============================================================================

process_maps <- function(baseline_path = NULL, metal_path = NULL) {
  
  # Inner helper: classify variants, bin into histogram, return plot data
  prepare_plot_data <- function(data, context, score_column = "score") {
    
    # Classify by HGVS notation suffix
    stop.is <- which(grepl("Ter$", data$hgvs_pro))   # nonsense: ends with 'Ter'
    syn.is  <- which(grepl("=$", data$hgvs_pro))      # synonymous: ends with '='
    miss.is <- setdiff(1:nrow(data), c(stop.is, syn.is))  # everything else = missense
    
    stp  <- na.omit(data[stop.is, score_column])
    syn  <- na.omit(data[syn.is, score_column])
    miss <- na.omit(data[miss.is, score_column])
    
    # Fraction of missense variants above 1.1 (potential gain-of-function)
    percent_miss_gt_1.1 <- if (length(miss) > 0) {
      sum(miss > 1.1) / length(miss) * 100
    } else { NA }
    
    medians <- data.frame(
      type = c("nonsense", "synonymous", "missense"),
      median = c(median(stp), median(syn), median(miss)),
      context = context
    )
    
    # Bin scores into fixed-width histogram bins (-2 to 2.5, width 0.1)
    stp_hist  <- hist(stp,  breaks = seq(-2, 2.5, 0.1), plot = FALSE)
    syn_hist  <- hist(syn,  breaks = seq(-2, 2.5, 0.1), plot = FALSE)
    miss_hist <- hist(miss, breaks = seq(-2, 2.5, 0.1), plot = FALSE)
    
    # Negate missense density for mirrored plot
    stp_df  <- data.frame(bin = stp_hist$mids,  density = stp_hist$density,  type = "nonsense")
    syn_df  <- data.frame(bin = syn_hist$mids,  density = syn_hist$density,  type = "synonymous")
    miss_df <- data.frame(bin = miss_hist$mids, density = -miss_hist$density, type = "missense")
    
    combined <- rbind(stp_df, syn_df, miss_df)
    combined$context <- context
    list(plot_data = combined, medians = medians, percent_miss_gt_1.1 = percent_miss_gt_1.1)
  }
  
  data_list    <- list()
  medians_list <- list()
  percentages  <- list()
  contexts     <- c("Baseline", "Metal")
  
  if (!is.null(baseline_path)) {
    baseline_map <- read_sheet("Baseline Map")
    baseline_results <- prepare_plot_data(baseline_map, "Baseline")
    data_list[["Baseline"]]   <- baseline_results$plot_data
    medians_list[["Baseline"]] <- baseline_results$medians
    percentages[["Baseline"]]  <- baseline_results$percent_miss_gt_1.1
  }
  
  if (!is.null(metal_path)) {
    metal_map <- read_sheet("Mercury Map")
    metal_results <- prepare_plot_data(metal_map, "Metal")
    data_list[["Metal"]]   <- metal_results$plot_data
    medians_list[["Metal"]] <- metal_results$medians
    percentages[["Metal"]]  <- metal_results$percent_miss_gt_1.1
  }
  
  print("Medians for each group:")
  medians_df <- do.call(rbind, medians_list)
  print(medians_df)
  
  print("Percentage of missense variants > 1.1:")
  percentages_df <- data.frame(
    context = names(percentages),
    percent_miss_gt_1.1 = unlist(percentages)
  )
  print(percentages_df)
  
  # Plot and save a PDF for each condition
  for (context in contexts) {
    if (!is.null(data_list[[context]])) {
      plot_data <- data_list[[context]]
      p <- ggplot(plot_data, aes(x = bin, y = density, fill = type)) +
        geom_bar(stat = "identity", position = "identity",
                 width = 0.1, color = "black", alpha = 0.4) +
        scale_fill_manual(values = c(
          "nonsense"   = "#C9364E",
          "synonymous" = "#003366",
          "missense"   = "grey"
        )) +
        labs(
          title = paste("Histogram of Variant Densities:", context),
          x = "Functional Score",
          y = "Density"
        ) +
        theme_minimal() +
        theme(
          axis.title = element_text(size = 14),
          axis.text  = element_text(size = 12),
          legend.position = "top"
        )
      print(p)
      
      filename <- paste0("~/Desktop/", context, "_Histogram.pdf")
      ggsave(filename, plot = p, width = 8, height = 6)
    }
  }
}

process_maps(
  baseline_path = "~/Desktop/Manuscript_Data.xlsx",
  metal_path    = "~/Desktop/Manuscript_Data.xlsx"
)
```


```{r, warning=FALSE, Consensus}
# =============================================================================
# CHUNK 2 — Consensus Circular Map
#
# Builds a polar (circular) visualisation of the full CPOX protein showing:
#   • Outer ring   : Baseline consensus (fraction of variants per score class)
#   • Middle ring  : Delta ring — shows variants that are APR (activity reduced
#                    in presence of mercury) or Improved under metal exposure
#   • Inner ring   : Metal map consensus
#   • Lollipops    : Clinical reference variants (HCP/HP = purple, benign = green)
#   • Secondary structure: helices (sine waves), strands (rectangles + arrowheads),
#                    loops (lines) fetched live from UniProt
#
# Score class boundaries (LOF → GOF):
#   lof1 ≤ 0.0 < lof3 ≤ 0.2 < lof4 ≤ 0.4 < lof5 ≤ 0.6 < lof6 ≤ 0.8
#   wt   ≤ 1.2 < gof1 ≤ 1.4 < gof2 ≤ 1.6 < gof3 ≤ 1.8 < gof4 ≤ 2.0
#   gof6  > 2.0
# =============================================================================

DATA_PATH <- "~/Desktop/Manuscript_Data.xlsx"

UNIPROT_ID <- "P36551"

class_levels <- c(
  "lof1","lof3","lof4","lof5","lof6",
  "wt",
  "gof1","gof2","gof3","gof4","gof6"
)

color_map <- c(
  lof1 = rgb(0.23, 0.37, 0.80),
  lof3 = rgb(0.48, 0.58, 0.87),
  lof4 = rgb(0.61, 0.69, 0.90),
  lof5 = rgb(0.74, 0.79, 0.93),
  lof6 = rgb(0.87, 0.89, 0.96),
  wt   = rgb(1, 1, 1),
  gof1 = rgb(0.96, 0.85, 0.85),
  gof2 = rgb(0.93, 0.71, 0.71),
  gof3 = rgb(0.90, 0.57, 0.57),
  gof4 = rgb(0.87, 0.43, 0.43),
  gof6 = rgb(0.80, 0.15, 0.15)
)

delta_colors <- c(
  turquoise  = "#2ED1D1",
  darkorange = "darkorange",
  white      = "white"
)

# ── Radial geometry parameters ────────────────────────────────────────────
ring_width        <- 0.45
delta_ring_height <- 0.03
ring_gap          <- 0.01

metal_inner_radius    <- 0.40
delta_inner_radius    <- metal_inner_radius + ring_width + ring_gap
baseline_inner_radius <- delta_inner_radius + delta_ring_height + ring_gap
structure_radius      <- baseline_inner_radius + ring_width + 0.15

lollipop_distance    <- 0.01
lollipop_height      <- 0.12
lollipop_base_radius <- structure_radius + lollipop_distance

gap_start <- 454
gap_width <- 20
gap_pos   <- (gap_start + 1):(gap_start + gap_width)

# ── Fetch UniProt secondary structure annotation ──────────────────────────
get_uniprot_secstruc <- function(uniprot_id) {
  js <- fromJSON(content(
    GET(paste0("https://rest.uniprot.org/uniprotkb/", uniprot_id, ".json")),
    "text"
  ))
  
  sec <- tibble(
    pos      = 1:nchar(js$sequence$value),
    secstruc = "other"
  )
  
  for (i in seq_len(nrow(js$features))) {
    f <- js$features[i, ]
    if (f$type %in% c("Helix", "Beta strand")) {
      s <- as.integer(f$location$start$value)
      e <- as.integer(f$location$end$value)
      sec$secstruc[s:e] <- ifelse(f$type == "Helix", "helix", "strand")
    }
  }
  sec
}

sec_df <- get_uniprot_secstruc(UNIPROT_ID)

# ── Build consensus rings ─────────────────────────────────────────────────
build_consensus <- function(path, sheet, r_inner, r_height, flip = FALSE) {
  df <- read_excel(path, sheet = sheet, skip = 1) %>%
    mutate(pos = as.integer(str_extract(hgvs_pro, "\\d+"))) %>%
    filter(!is.na(pos), !is.na(score)) %>%
    mutate(class = case_when(
      score <= 0.0 ~ "lof1", score <= 0.2 ~ "lof3", score <= 0.4 ~ "lof4",
      score <= 0.6 ~ "lof5", score <= 0.8 ~ "lof6", score <= 1.2 ~ "wt",
      score <= 1.4 ~ "gof1", score <= 1.6 ~ "gof2", score <= 1.8 ~ "gof3",
      score <= 2.0 ~ "gof4", TRUE ~ "gof6"
    ))
  
  out <- df %>%
    count(pos, class, name = "n") %>%
    mutate(class = factor(class, levels = class_levels)) %>%
    arrange(pos, class) %>%
    group_by(pos) %>%
    mutate(
      frac = n / sum(n),
      ymin = if (!flip) r_inner + lag(cumsum(frac), default = 0) * r_height
      else r_inner + r_height - cumsum(frac) * r_height,
      ymax = if (!flip) r_inner + cumsum(frac) * r_height
      else r_inner + r_height - lag(cumsum(frac), default = 0) * r_height
    ) %>%
    ungroup()
  
  bind_rows(
    out,
    tibble(
      pos   = gap_pos,
      class = factor(NA, levels = class_levels),
      ymin  = r_inner,
      ymax  = r_inner + r_height
    )
  )
}

# ── Build delta ring ──────────────────────────────────────────────────────
build_delta_ring <- function(delta_path, delta_sheet, baseline_path, baseline_sheet, r_inner, r_height) {
  
  delta <- read_excel(delta_path, sheet = delta_sheet, skip = 1) %>%
    mutate(
      deltaScore  = as.numeric(deltaScore),
      deltaSE     = as.numeric(deltaSE),
      pos         = as.integer(str_extract(hgvs_pro, "\\d+")),
      significant = case_when(
      significant == "TRUE"  ~ TRUE,
      significant == "FALSE" ~ FALSE,
      TRUE                   ~ NA
)    ) %>%
    filter(!is.na(pos), is.finite(deltaScore), is.finite(deltaSE), !is.na(significant))
  setDT(delta)
  
  baseline_map <- read_excel(baseline_path, sheet = baseline_sheet, skip = 1)
  setDT(baseline_map)
  colnames(baseline_map)[1:4] <- c("hgvs_pro", "baseline_score", "se", "df")
  
  baseline_map[, baseline_lower := baseline_score + se]
  baseline_stop <- grepl("Ter$", baseline_map$hgvs_pro)
  baseline_syn  <- grepl("=$",   baseline_map$hgvs_pro)
  baseline_miss <- !(baseline_stop | baseline_syn)
  
  baseline_miss_tolerated <- baseline_map[
    baseline_miss & baseline_lower >= 0.5
  ]
  
  delta[, Impact := "neutral"]
  APR_set      <- delta[significant == TRUE  & deltaScore < 0 & hgvs_pro %in% baseline_miss_tolerated$hgvs_pro, hgvs_pro]
  Improved_set <- delta[significant == TRUE  & deltaScore > 0 & hgvs_pro %in% baseline_miss_tolerated$hgvs_pro, hgvs_pro]
  known_set    <- delta[significant == FALSE & hgvs_pro %in% baseline_miss_tolerated$hgvs_pro, hgvs_pro]
  
  delta[hgvs_pro %in% APR_set,      Impact := "APR"]
  delta[hgvs_pro %in% Improved_set, Impact := "improved"]
  delta[hgvs_pro %in% known_set,    Impact := "known"]
  
  delta_pos <- delta %>%
    group_by(pos, Impact) %>%
    summarise(n = n(), .groups = "drop") %>%
    group_by(pos) %>%
    mutate(
      max_n    = max(n),
      n_at_max = sum(n == max_n)
    ) %>%
    slice_max(n, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(
      delta_bin = case_when(
        Impact == "APR"      & n_at_max == 1 ~ "turquoise",
        Impact == "improved" & n_at_max == 1 ~ "darkorange",
        TRUE                                 ~ "white"
      ),
      ymin = r_inner,
      ymax = r_inner + r_height
    ) %>%
    select(pos, delta_bin, ymin, ymax)
  
  bind_rows(
    delta_pos,
    tibble(pos = gap_pos, delta_bin = "white", ymin = r_inner, ymax = r_inner + r_height)
  )
}

# ── Secondary structure elements ──────────────────────────────────────────
make_ss_elements <- function(sec_df, r) {
  sec_df <- bind_rows(sec_df, tibble(pos = gap_pos, secstruc = "gap"))
  helices <- strands <- arrowheads <- loops <- tibble()
  
  i <- 1
  while (i <= nrow(sec_df)) {
    t <- sec_df$secstruc[i]
    s <- sec_df$pos[i]
    while (i < nrow(sec_df) && sec_df$secstruc[i + 1] == t) i <- i + 1
    e <- sec_df$pos[i]
    
    if (t == "helix") {
      x <- seq(s, e, length.out = 80)
      y <- r + 0.08 * sin(seq(0, 2 * pi, length.out = 80))
      helices <- bind_rows(helices, tibble(x, y, group = s))
    } else if (t == "strand") {
      h <- 0.12
      strands <- bind_rows(strands, tibble(
        x = c(s, e, e, s), y = c(r - h/2, r - h/2, r + h/2, r + h/2), group = s
      ))
      arrowheads <- bind_rows(arrowheads, tibble(
        x = c(e, e + 3, e), y = c(r - h/2, r, r + h/2), group = s
      ))
    } else {
      loops <- bind_rows(loops, tibble(x = c(s, e), y = c(r, r), group = s))
    }
    i <- i + 1
  }
  
  list(helices, strands, arrowheads, loops)
}

ss <- make_ss_elements(sec_df, structure_radius)

# ── Clinical reference lollipops ──────────────────────────────────────────
ref <- read_excel(DATA_PATH, sheet = "ReferenceSet", skip = 1) %>%
  mutate(
    pos = as.integer(str_extract(hgvs_pro, "\\d+")),
    col = case_when(
      disease_type %in% c("HCP", "HP") ~ "#AF22DD",
      disease_type == "VUS"            ~ "grey50",
      disease_type == "none"           ~ "#50DD22",
      TRUE                             ~ NA_character_
    )
  ) %>%
  filter(!is.na(pos), !is.na(col))

# ── Build all rings ───────────────────────────────────────────────────────
metal <- build_consensus(
  DATA_PATH, "Mercury Map",
  metal_inner_radius, ring_width, flip = TRUE
)

delta <- build_delta_ring(
  DATA_PATH, "Delta Map",
  DATA_PATH, "Baseline Map",
  delta_inner_radius, delta_ring_height
)

base <- build_consensus(
  DATA_PATH, "Baseline Map",
  baseline_inner_radius, ring_width
)

axis_breaks <- c(1, 100, 200, 300, 400)

# ── Shared plot builder ───────────────────────────────────────────────────
build_plot <- function() {
  ggplot() +
    geom_rect(
      data = metal,
      aes(xmin = pos - 0.5, xmax = pos + 0.5, ymin = ymin, ymax = ymax, fill = class),
      color = "grey80", linewidth = 0.2
    ) +
    scale_fill_manual(values = color_map, na.value = "white") +
    new_scale_fill() +
    geom_rect(
      data = delta,
      aes(xmin = pos - 0.5, xmax = pos + 0.5, ymin = ymin, ymax = ymax, fill = delta_bin),
      color = "grey80", linewidth = 0.2
    ) +
    scale_fill_manual(values = delta_colors, na.value = "white") +
    new_scale_fill() +
    geom_rect(
      data = base,
      aes(xmin = pos - 0.5, xmax = pos + 0.5, ymin = ymin, ymax = ymax, fill = class),
      color = "grey80", linewidth = 0.2
    ) +
    scale_fill_manual(values = color_map, na.value = "white") +
    geom_segment(
      data = ref,
      aes(x = pos, xend = pos,
          y = lollipop_base_radius,
          yend = lollipop_base_radius + lollipop_height),
      color = ref$col, linewidth = 0.5
    ) +
    geom_point(
      data = ref,
      aes(x = pos, y = lollipop_base_radius + lollipop_height),
      color = ref$col, size = 1
    ) +
    geom_path(data = ss[[4]], aes(x, y, group = group), color = "grey40", linewidth = 2)  +
    geom_polygon(data = ss[[2]], aes(x, y, group = group), fill = "black")                +
    geom_polygon(data = ss[[3]], aes(x, y, group = group), fill = "black")                +
    geom_line(data = ss[[1]], aes(x, y, group = group), color = "black", linewidth = 0.3) +
    coord_polar(start = -pi/2) +
    ylim(0, lollipop_base_radius + lollipop_height + 0.1) +
    scale_x_continuous(breaks = axis_breaks, labels = axis_breaks) +
    theme(
      axis.text.x        = element_text(color = "black", size = 10),
      panel.background   = element_rect(fill = "transparent", color = NA),
      plot.background    = element_rect(fill = "transparent", color = NA),
      panel.grid.major.x = element_line(color = "black", size = 0.2),
      axis.ticks         = element_line(color = "black")
    )
}

# ── Render interactive preview ────────────────────────────────────────────
build_plot()

# ── Export to PDF ─────────────────────────────────────────────────────────
pdf(file = "~/Desktop/Fig1_Maps.pdf", width = 8, height = 8, useDingbats = FALSE)
print(build_plot())
dev.off()

# ── Count residue positions where the dominant delta impact is APR (teal) ──
baseline_map <- read_excel(DATA_PATH, sheet = "Baseline Map", skip = 1)
colnames(baseline_map)[1:4] <- c("hgvs_pro", "baseline_score", "se", "df")
baseline_map <- baseline_map %>% mutate(baseline_lower = baseline_score + se)

baseline_miss_tolerated <- baseline_map %>%
  filter(!grepl("Ter$", hgvs_pro), !grepl("=$", hgvs_pro), baseline_lower >= 0.5)

teal_positions <- delta$pos[delta$delta_bin == "turquoise"]

delta_raw <- read_excel(DATA_PATH, sheet = "Delta Map", skip = 1) %>%
  mutate(
    deltaScore  = as.numeric(deltaScore),
    deltaSE     = as.numeric(deltaSE),
    pos         = as.integer(str_extract(hgvs_pro, "\\d+")),
    significant = case_when(
    significant == "TRUE"  ~ TRUE,
    significant == "FALSE" ~ FALSE,
  TRUE                   ~ NA
)  ) %>%
  filter(!is.na(significant))

delta_raw %>%
  filter(
    pos %in% teal_positions,
    significant == TRUE,
    deltaScore < 0,
    hgvs_pro %in% baseline_miss_tolerated$hgvs_pro
  ) %>%
  summarise(n_positions = n_distinct(pos), n_variants = n())
```


```{r, Population Level Assessment, eval=FALSE}
# =============================================================================
# CHUNK 3 — Population-Level Assessment
#
# Tests whether the DMS functional scores are depleted for deleterious variants
# in the human population, as expected under purifying selection.
#
# Method: log odds ratio comparing the observed frequency of deleterious /
# neutral DMS variants in population databases vs. their expected frequency
# under no selection. Fisher's exact test provides 95% CIs.
#
# Population sources:
#   • UK Biobank — allele frequencies from the UK Biobank exome release
#   • gnomAD v4  — population allele frequencies (gnomAD missense variants)
# =============================================================================

# ── Log odds ratio function ────────────────────────────────────────────────
calculate_log_odds_ci <- function(CPOX_data, human_pop_data,
                                  deleterious_threshold, neutral_threshold) {
  
  deleterious_variants <- CPOX_data %>%
    filter(score < deleterious_threshold) %>%
    select(hgvs_pro)
  
  neutral_variants <- CPOX_data %>%
    filter(score > neutral_threshold) %>%
    select(hgvs_pro)
  
  # Observed: how many DMS variants appear in the population dataset
  observed_del <- sum(deleterious_variants$hgvs_pro %in% human_pop_data$hgvs_pro)
  total_del    <- nrow(deleterious_variants)
  expected_del <- (nrow(human_pop_data) / nrow(CPOX_data)) * total_del
  
  observed_neu <- sum(neutral_variants$hgvs_pro %in% human_pop_data$hgvs_pro)
  total_neu    <- nrow(neutral_variants)
  expected_neu <- (nrow(human_pop_data) / nrow(CPOX_data)) * total_neu
  
  # 2×2 contingency tables: observed vs. expected
  del_table <- matrix(
    c(observed_del, total_del - observed_del,
      expected_del, total_del - expected_del),
    nrow = 2, byrow = TRUE
  )
  
  neu_table <- matrix(
    c(observed_neu, total_neu - observed_neu,
      expected_neu, total_neu - expected_neu),
    nrow = 2, byrow = TRUE
  )
  
  log_odds_del <- log((del_table[1,1] * del_table[2,2]) / (del_table[1,2] * del_table[2,1]))
  log_odds_neu <- log((neu_table[1,1] * neu_table[2,2]) / (neu_table[1,2] * neu_table[2,1]))
  
  fisher_del <- fisher.test(del_table)
  fisher_neu <- fisher.test(neu_table)
  
  list(
    log_odds_del = log_odds_del,
    log_odds_neu = log_odds_neu,
    ci_del       = log(fisher_del$conf.int),
    ci_neu       = log(fisher_neu$conf.int)
  )
}

# ── UK Biobank ────────────────────────────────────────────────────────────
# NOTE: UKBiobank sheet must be present in Manuscript_Data.xlsx
UKBiobank <- tryCatch({
  df <- read_sheet("UKBiobank")
  if (nrow(df) == 0) stop("empty")
  # Build HGVS protein notation from single-letter amino acid columns
  df$hgvs_pro <- paste0(
    substr(df$Amino_acids, 1, 1),
    df$Protein_position,
    substr(df$Amino_acids, nchar(df$Amino_acids), nchar(df$Amino_acids))
  )
  # Convert single-letter amino acid codes to three-letter (e.g. "A123V" → "Ala123Val")
  df$hgvs_pro <- mgsub(
    df$hgvs_pro,
    pattern     = c("A","R","N","D","C","Q","E","G","H","I","L","K","M","F","P","O","S","U","T","W","Y","V"),
    replacement = c("Ala","Arg","Asn","Asp","Cys","Gln","Glu","Gly","His","Ile","Leu","Lys","Met","Phe","Pro","Pyl","Ser","Sec","Thr","Trp","Tyr","Val")
  )
  df$hgvs_pro <- paste0("p.", df$hgvs_pro)
  df <- df %>%
    distinct(hgvs_pro, .keep_all = TRUE) %>%
    filter(!grepl("X|\\*", hgvs_pro)) %>%
    select(hgvs_pro, UKB_AF)
  colnames(df)[2] <- "frequency"
  df
}, error = function(e) {
  warning("UKBiobank sheet missing or empty in Manuscript_Data.xlsx — skipping. ",
          "Add UKBiobank.csv data to that sheet (info row in row 1, headers in row 2).")
  data.frame(hgvs_pro = character(), frequency = numeric())
})

# ── gnomAD v4 ─────────────────────────────────────────────────────────────
# NOTE: gnomADv4_Data sheet must be present in Manuscript_Data.xlsx
gnomad <- tryCatch({
  df <- read_sheet("gnomADv4_Data")
  if (nrow(df) == 0) stop("empty")
  colnames(df)[12] <- "hgvs_pro"
  colnames(df)[14] <- "type"
  colnames(df)[20] <- "frequency"
  df %>%
    filter(type == "missense_variant") %>%
    select(hgvs_pro, frequency) %>%
    distinct(hgvs_pro, .keep_all = TRUE)
}, error = function(e) {
  warning("gnomADv4_Data sheet missing or empty in Manuscript_Data.xlsx — skipping. ",
          "Add gnomADv4_Data.csv data to that sheet (info row in row 1, headers in row 2).")
  data.frame(hgvs_pro = character(), frequency = numeric())
})

# ── Combine population datasets ────────────────────────────────────────────
HumanPop_Frequency <- bind_rows(UKBiobank, gnomad) %>%
  distinct(hgvs_pro, .keep_all = TRUE)

cat("Total unique variants in combined population dataset:", nrow(HumanPop_Frequency), "\n")

# ── Calculate log odds ─────────────────────────────────────────────────────
baseline <- read_sheet("Baseline Map")

deleterious_threshold <- 0.5
neutral_threshold     <- 0.5

log_odds <- calculate_log_odds_ci(baseline, HumanPop_Frequency,
                                   deleterious_threshold, neutral_threshold)

log_odds_data <- data.frame(
  category = rep("baseline", 2),
  Type     = c("deleterious", "neutral"),
  Log_Odds = c(log_odds$log_odds_del, log_odds$log_odds_neu),
  CI_Lower = c(log_odds$ci_del[1],   log_odds$ci_neu[1]),
  CI_Upper = c(log_odds$ci_del[2],   log_odds$ci_neu[2])
)

# Point-range plot (log odds with 95% CI)
pointRangeOdds <- ggplot(log_odds_data, aes(x = Type, y = Log_Odds)) +
  geom_pointrange(
    aes(ymin = CI_Lower, ymax = CI_Upper, color = category),
    position = position_dodge(0.9),
    size = 0.5
  ) +
  scale_color_manual(values = c("baseline" = "black")) +
  labs(x = "Variant Type", y = "Log Odds Ratio") +
  theme_minimal(base_size = 15) +
  coord_flip()

print(pointRangeOdds)
```


```{r, warning=FALSE, Comparison of Baseline and Metal Map Scores}
# =============================================================================
# CHUNK 4 — Baseline vs. Metal Map Correlation
#
# Tests the global agreement between the baseline and metal-condition DMS maps.
# High correlation indicates that relative variant effects are preserved under
# metal exposure; divergence reveals condition-specific responses.
#
# Pearson r is used (linear relationship expected between scaled scores).
# Dashed orange lines at score = 0.5 mark the LOF/functional boundary.
# =============================================================================

baseline_map <- read_sheet("Baseline Map")
metal_map    <- read_sheet("Mercury Map")

names(baseline_map)[names(baseline_map) == "score"] <- "baseline"
names(metal_map)[names(metal_map)       == "score"] <- "metal"

merged_maps     <- merge(baseline_map, metal_map, by = "hgvs_pro")
merged_maps$pos <- as.numeric(gsub("\\D", "", merged_maps$hgvs_pro))

cor_pear <- cor.test(merged_maps$baseline, merged_maps$metal, method = "pearson")
cat(sprintf("Pearson r = %.2f, p = %.3g\n", cor_pear$estimate, cor_pear$p.value))
r_value_spear <- cor(merged_maps$baseline, merged_maps$metal, method = "spearman")

maps_corr <- ggplot(merged_maps, aes(x = baseline, y = metal)) +
  geom_point(alpha = 0.15) +
  geom_smooth(method = "lm", se = FALSE, color = "turquoise", linewidth = 0.8) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "darkorange") +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "darkorange") +
  labs(x = "Baseline CPOX Map", y = "Conditional CPOX Map \n (Metal;Hg)") +
  theme_minimal(base_size = 15) +
  annotate(
    "text",
    x = max(merged_maps$baseline),
    y = min(merged_maps$metal),
    label = paste("R = ", round(cor_pear$estimate, 2)),
    hjust = 1
  )

pdf("~/Desktop/BetweenMapCorrelation.pdf", width = 8, height = 6)
print(maps_corr)
dev.off()
```


```{r, warning=FALSE, Examining Delta Scores}
# =============================================================================
# CHUNK 5 — Delta Score Analysis
#
# The delta score captures the change in functional fitness between the
# baseline and metal (Hg) conditions: delta = baseline_logPhi – metal_logPhi
# Positive delta → variant tolerates baseline but not metal (APR)
# Negative delta → variant is improved by metal exposure
#
# This chunk:
#   1. Tests whether there is a systematic shift using a paired Wilcoxon test
#   2. Plots a mirrored bar histogram: controls (synonymous + nonsense) above
#      zero, missense below zero
#   3. Plots the full distribution of all delta scores
# =============================================================================

delta_map <- read_sheet("Delta Map") %>% na.omit()

# Overall median shift between conditions
median_diff <- median(
  delta_map$logPhi_weighted_mean_metal - delta_map$logPhi_weighted_mean_baseline,
  na.rm = TRUE
)

# Paired Wilcoxon: tests whether the two conditions differ per variant
wilcox_res <- wilcox.test(
  delta_map$logPhi_weighted_mean_baseline,
  delta_map$logPhi_weighted_mean_metal,
  paired      = TRUE,
  alternative = "two.sided",
  exact       = FALSE
)

cat("Median difference (metal - baseline):", median_diff, "\n")
print(wilcox_res)

process_delta <- function(delta_path) {
  
  data <- read_sheet("Delta Map") %>% na.omit()
  
  stop_idx  <- grepl("Ter$", data$hgvs_pro)
  syn_idx   <- grepl("=$",   data$hgvs_pro)
  miss_idx  <- !(stop_idx | syn_idx)
  
  cntl_scores <- data$deltaScore[stop_idx | syn_idx]
  miss_scores <- data$deltaScore[miss_idx]
  
  # ── Plot 1: Mirrored density histogram ────────────────────────────────
  hist_breaks <- seq(-3, 3, by = 0.1)
  cntl_hist   <- hist(cntl_scores, breaks = hist_breaks, plot = FALSE)
  miss_hist   <- hist(miss_scores, breaks = hist_breaks, plot = FALSE)
  
  plot_df <- rbind(
    data.frame(bin = cntl_hist$mids, density = cntl_hist$density,   type = "cntl"),
    data.frame(bin = miss_hist$mids, density = -miss_hist$density,  type = "missense")
  )
  
  Grouped_Delta <- ggplot(plot_df, aes(x = bin, y = density, fill = type)) +
    geom_col(width = 0.095) +
    scale_fill_manual(values = c("cntl" = "black", "missense" = "grey60")) +
    labs(
      title = "Delta Score Density: Controls vs Missense",
      x = "Delta Score",
      y = "Density"
    ) +
    theme_minimal() +
    theme(axis.title = element_text(size = 14), axis.text = element_text(size = 12),
          legend.position = "top")
  
  pdf("~/Desktop/group_Delta.pdf", width = 8, height = 6)
  print(Grouped_Delta)
  dev.off()
  
  # ── Plot 2: Full delta score distribution ──────────────────────────────
  complete_Delta_dis <- ggplot(
    data.frame(deltaScore = data$deltaScore),
    aes(x = deltaScore)
  ) +
    geom_histogram(
      aes(y = after_stat(density)),
      binwidth = 0.05, fill = "grey40", color = "black", alpha = 0.6
    ) +
    geom_density(linewidth = 1.2) +
    labs(title = "Distribution of All Delta Scores", x = "Delta Score", y = "Density") +
    theme_minimal() +
    theme(title = element_text(size = 14, face = "bold"),
          axis.title = element_text(size = 14), axis.text = element_text(size = 12))
  
  pdf("~/Desktop/complete_Delta.pdf", width = 8, height = 6)
  print(complete_Delta_dis)
  dev.off()
}

process_delta("~/Desktop/Manuscript_Data.xlsx")  # reads sheet "Delta Map"
```


```{r, warning=FALSE, Comparison of Baseline & Metal Map Scores with Predicted ΔΔG Scores}
# =============================================================================
# CHUNK 6 — Functional Scores vs. Predicted ΔΔG
#
# Compares DMS scores (baseline and metal) to in silico thermodynamic
# stability predictions (ΔΔG) based on the CPOX crystal structure (PDB: 2AEX).
#
# Workflow:
#   1. Convert ΔΔG variant notation to HGVS
#   2. Merge with DMS scores
#   3. Compute 15-aa rolling means per position
#   4. Plot rolling mean traces for baseline, metal, and ΔΔG
#   5. Pearson correlation with BH-adjusted p-values
#   6. Scatter plots with r annotation
# =============================================================================

ddG_scores <- read_sheet("ddG Values")
#ddG_scores <- subset(ddG_scores, CHAIN != "")

# Single-letter → three-letter amino acid dictionary
aa_dict <- c("A"="Ala","R"="Arg","N"="Asn","D"="Asp","C"="Cys","Q"="Gln",
             "E"="Glu","G"="Gly","H"="His","I"="Ile","L"="Leu","K"="Lys",
             "M"="Met","F"="Phe","P"="Pro","S"="Ser","T"="Thr","W"="Trp",
             "Y"="Tyr","V"="Val")



# Convert "A123V" → "p.Ala123Val"
replace_aa <- function(s) {
  for (aa in names(aa_dict)) s <- gsub(aa, aa_dict[aa], s)
  return(paste0("p.", s))
}

ddG_scores$VARIANT <- replace_aa(ddG_scores$VARIANT)
ddG_scores <- ddG_scores[, c(1, 3)]
colnames(ddG_scores) <- c("hgvs_pro", "ddG")

ddG_scores$hgvs_pro <- ddG_scores$hgvs_pro %>%
  gsub("Glyln", "Gln", .) %>%
  gsub("Glylu", "Glu", .) %>%
  gsub("Prohe", "Phe", .)

# Load and rename map columns
baseline_map <- read_sheet("Baseline Map")
colnames(baseline_map)[2:4] <- c("baseline", "baselineSE", "baselineDF")

metal_map <- read_sheet("Mercury Map")
setDT(metal_map)
colnames(metal_map)[2:4] <- c("metal", "metalSE", "metalDF")

# Merge and remove synonymous variants (regex: same AA before and after position)
Stability <- ddG_scores |>
  merge(baseline_map, by = "hgvs_pro") |>
  merge(metal_map,    by = "hgvs_pro") |>
  subset(!grepl("(\\w\\d+)(\\1)", hgvs_pro)) |>
  unique()

Stability$pos <- as.numeric(gsub("\\D", "", Stability$hgvs_pro))

# 15-aa rolling mean smoothing for positional trends
Rolling_Stability <- Stability |>
  dplyr::group_by(pos) |>
  dplyr::summarize(
    mean_baseline = mean(baseline,       na.rm = TRUE),
    mean_metal    = mean(metal,          na.rm = TRUE),
    mean_ddG      = mean(as.integer(ddG), na.rm = TRUE),
    .groups       = "drop"
  ) |>
  dplyr::mutate(
    rolling_baseline = zoo::rollmean(mean_baseline, k = 15, fill = NA, align = "center"),
    rolling_metal    = zoo::rollmean(mean_metal,    k = 15, fill = NA, align = "center"),
    rolling_ddG      = zoo::rollmean(mean_ddG,      k = 15, fill = NA, align = "center")
  )

# Melt to long format for faceted plotting
Rolling_Stability_long <- reshape2::melt(
  Rolling_Stability[, c("pos","rolling_baseline","rolling_metal","rolling_ddG")],
  id.vars       = "pos",
  variable.name = "Metric",
  value.name    = "Score"
)

# Rolling mean line plot per metric
plot_metric <- function(data, metric_name, color, ylab) {
  ggplot(data[data$Metric == metric_name, ], aes(x = pos, y = Score)) +
    geom_line(size = 0.5, color = color) +
    labs(x = "Amino Acid Position", y = ylab) +
    geom_vline(xintercept = seq(0, 450, by = 50), color = "grey", linetype = "dotted") +
    theme_minimal() +
    theme(panel.grid.major.x = element_line(color = "grey", linetype = "dotted"),
          axis.text.x = element_text(angle = 0, hjust = 1)) +
    scale_x_continuous(limits = range(data$pos, na.rm = TRUE))
}

p1 <- plot_metric(Rolling_Stability_long, "rolling_baseline", "#11EE89", "Baseline Score\n(Moving Avg, 15 AA)")
p2 <- plot_metric(Rolling_Stability_long, "rolling_metal",    "#EE1176", "Metal Score\n(Moving Avg, 15 AA)")
p3 <- plot_metric(Rolling_Stability_long, "rolling_ddG",      "black",   "Predicted ΔΔG\n(Moving Avg, 15 AA)")
gridExtra::grid.arrange(p1, p2, p3, ncol = 1)

# Pearson correlations between functional scores and ΔΔG
cor_baseline <- cor.test(Stability$baseline, Stability$ddG, method = "pearson")
cor_metal    <- cor.test(Stability$metal,    Stability$ddG, method = "pearson")

# BH-adjusted p-values (two tests)
raw_pvals <- c(cor_baseline$p.value, cor_metal$p.value)
padj      <- p.adjust(raw_pvals, method = "BH")
names(padj) <- c("Baseline vs ΔΔG", "Metal vs ΔΔG")

cat(sprintf("Baseline vs ΔΔG: Pearson r = %.2f, p = %.3g, padj = %.3g\n",
            cor_baseline$estimate, cor_baseline$p.value, padj[1]))
cat(sprintf("Metal vs ΔΔG:    Pearson r = %.2f, p = %.3g, padj = %.3g\n",
            cor_metal$estimate, cor_metal$p.value, padj[2]))

# Scatter plot helper
plot_scatter <- function(x, y, cor_result, padj_value, title, color) {
  ggplot(Stability, aes_string(x = x, y = y)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", color = color) +
    annotate("text",
             x     = max(Stability[[x]], na.rm = TRUE),
             y     = min(Stability[[y]], na.rm = TRUE),
             label = sprintf("r = %.2f\np = %.3g", cor_result$estimate, cor_result$p.value),
             hjust = 1, vjust = 0, size = 4) +
    labs(x = x, y = "Predicted ΔΔG Score", title = title) +
    theme_minimal() +
    theme(panel.grid.major = element_line(color = "grey", linetype = "dotted"),
          axis.text.x = element_text(angle = 0),
          axis.text.y = element_text(angle = 0))
}

scatter1 <- plot_scatter("baseline", "ddG", cor_baseline, padj[1], "Baseline Score vs. ΔΔG", "turquoise")
scatter2 <- plot_scatter("metal",    "ddG", cor_metal,    padj[2], "Metal Score vs. ΔΔG",    "purple")

pdf("~/Desktop/ddG&MapCorrelation.pdf",  width = 8, height = 6); print(scatter1); dev.off()
pdf("~/Desktop/ddG&MapCorrelation2.pdf", width = 8, height = 6); print(scatter2); dev.off()
```


```{r, warning=FALSE, Comparison of Baseline & Metal Map Scores with Patient Enzymatic Activity}
# =============================================================================
# CHUNK 7 — Correlation with Patient Enzymatic Activity
#
# Tests whether the DMS functional score predicts residual CPOX enzymatic
# activity measured in HCP patients. Activity is expressed as a percentage
# of the lower reference value.
#
# Both Pearson and Spearman correlations are reported. Median activity per
# variant is used (multiple patients may carry the same variant).
# =============================================================================

process_activity_correlation <- function(baseline_path = NULL,
                                          metal_path    = NULL,
                                          activity_path = NULL) {
  
  # Inner helper: merge map scores with activity data and aggregate per variant
  prepare_correlation_data <- function(map_data, activity_data, score_column) {
    merged <- merge(activity_data, map_data, by = "hgvs_pro")
    # Filter to rows where HCP patient enzyme activity is present
    merged <- merged[!is.na(merged$hcp_enzyme_activity), ]
    merged <- merged[, c("hgvs_pro", "pct_activity", score_column)]
    
    # Median activity per variant (collapses multiple patient entries)
    median_activity <- aggregate(
      pct_activity ~ hgvs_pro,
      data = merged, FUN = median, na.rm = TRUE
    )
    colnames(median_activity) <- c("hgvs_pro", "median_activity")
    
    out <- merge(map_data, median_activity, by = "hgvs_pro")
    out[complete.cases(out), ]
  }
  
  if (!is.null(activity_path)) {
    activity_data <- read_sheet("Enzymatic Activity")
    if (nrow(activity_data) == 0 || ncol(activity_data) == 0) {
      warning("Enzymatic Activity sheet is empty in Manuscript_Data.xlsx. ",
              "Please paste the CPOX_activities.csv data into that sheet ",
              "(with a generic info row in row 1 and column headers in row 2). ",
              "Skipping activity correlation.")
      return(invisible(NULL))
    }
    # Rename by position: col names contain newlines and special characters
    # Col 4 = Amino acid change (hgvs_pro)
    # Col 5 = HCP patients Enzyme activity
    # Col 8 = % of activity compared to the lower reference value
    colnames(activity_data)[c(4, 5, 8)] <- c("hgvs_pro", "hcp_enzyme_activity", "pct_activity")
    activity_data$hgvs_pro            <- as.character(activity_data$hgvs_pro)
    activity_data$hcp_enzyme_activity <- suppressWarnings(as.numeric(activity_data$hcp_enzyme_activity))
    activity_data$pct_activity        <- suppressWarnings(as.numeric(activity_data$pct_activity))
  } else {
    stop("Activity data required.")
  }
  
  # ── BASELINE ──────────────────────────────────────────────────────────
  if (!is.null(baseline_path)) {
    baseline_map <- read_sheet("Baseline Map")
    colnames(baseline_map)[2] <- "baseline_score"
    baseline_map <- baseline_map[, c("hgvs_pro", "baseline_score")]
    
    baseline_data <- prepare_correlation_data(baseline_map, activity_data, "baseline_score")
    
    spearman_baseline <- cor.test(baseline_data$median_activity, baseline_data$baseline_score, method = "spearman")
    pearson_baseline  <- cor.test(baseline_data$median_activity, baseline_data$baseline_score, method = "pearson")
    
    print(paste("Spearman (Baseline):", round(spearman_baseline$estimate, 2),
                "P =", signif(spearman_baseline$p.value, 3)))
    print(paste("Pearson (Baseline):",  round(pearson_baseline$estimate,  2),
                "P =", signif(pearson_baseline$p.value,  3)))
    
    pdf("~/Desktop/Baseline_Activity.pdf", width = 8, height = 6)
    plot(baseline_data$median_activity, baseline_data$baseline_score,
         xlab = "Activity (% of lower reference value)",
         ylab = "Baseline Score",
         main = paste("Spearman r =", round(spearman_baseline$estimate, 2),
                      "p =", signif(spearman_baseline$p.value, 2),
                      "\nPearson r =", round(pearson_baseline$estimate, 2),
                      "p =", signif(pearson_baseline$p.value, 2)),
         pch = 19, col = "grey40")
    abline(lm(baseline_score ~ median_activity, data = baseline_data), lwd = 2)
    dev.off()
  }
  
  # ── METAL ─────────────────────────────────────────────────────────────
  if (!is.null(metal_path)) {
    metal_map <- read_sheet("Mercury Map")
    colnames(metal_map)[2] <- "metal_score"
    metal_map <- metal_map[, c("hgvs_pro", "metal_score")]
    
    metal_data <- prepare_correlation_data(metal_map, activity_data, "metal_score")
    
    spearman_metal <- cor.test(metal_data$median_activity, metal_data$metal_score, method = "spearman")
    pearson_metal  <- cor.test(metal_data$median_activity, metal_data$metal_score, method = "pearson")
    
    print(paste("Spearman (Metal):", round(spearman_metal$estimate, 2),
                "P =", signif(spearman_metal$p.value, 3)))
    print(paste("Pearson (Metal):",  round(pearson_metal$estimate,  2),
                "P =", signif(pearson_metal$p.value,  3)))
    
    pdf("~/Desktop/Metal_Activity.pdf", width = 8, height = 6)
    plot(metal_data$median_activity, metal_data$metal_score,
         xlab = "Activity (% of lower reference value)",
         ylab = "Metal Score",
         main = paste("Spearman r =", round(spearman_metal$estimate, 2),
                      "p =", signif(spearman_metal$p.value, 2),
                      "\nPearson r =", round(pearson_metal$estimate, 2),
                      "p =", signif(pearson_metal$p.value, 2)),
         pch = 19, col = "grey40")
    abline(lm(metal_score ~ median_activity, data = metal_data), lwd = 2)
    dev.off()
  }
}

process_activity_correlation(
  baseline_path = "~/Desktop/Manuscript_Data.xlsx",
  metal_path    = "~/Desktop/Manuscript_Data.xlsx",
  activity_path = "~/Desktop/Manuscript_Data.xlsx"
)
```


```{r, warning=FALSE, Hypercomplementing variant scores: Baseline Map vs AlphaMissense}
# =============================================================================
# CHUNK 8 — Hypercomplementing Variants vs. AlphaMissense
#
# Hypercomplementing variants are defined as those with baseline score > 1.2
# (above the WT reference range). These may represent gain-of-function or
# hyperstable variants.
#
# This chunk compares the DMS baseline score against AlphaMissense predictions
# for this subset. AlphaMissense scores are rescaled to [0,1] for direct
# comparison. Expected: AlphaMissense should score these variants as
# damaging (high score) while DMS shows elevated function.
# =============================================================================

baseline_file <- read_sheet("Baseline Map")
setDT(baseline_file)
colnames(baseline_file)[2] <- "baseline_map"
baseline_file <- baseline_file[, .(hgvs_pro, baseline_map)]

AM_Data <- read_sheet("AlphaMissense Values")
setDT(AM_Data)
colnames(AM_Data)[1:2] <- c("hgvs_pro", "AM")

# Keep only hypercomplementing variants (score > 1.2)
Hyper_Data <- merge(AM_Data, baseline_file, by = "hgvs_pro")
Hyper_Data <- Hyper_Data[baseline_map > 1.2]
Hyper_Data <- na.omit(Hyper_Data)

# Rescale AlphaMissense to [0,1] for comparable y-axis with baseline
rescaled_AM  <- (Hyper_Data$AM - min(Hyper_Data$AM)) / (max(Hyper_Data$AM) - min(Hyper_Data$AM))
rescaled_Map <- Hyper_Data$baseline_map

Map_Preditor_Hyper_Scores <- data.table(
  AlphaMissense = rescaled_AM,
  BaselineMap   = rescaled_Map
)

Map_Preditor_Hyper_Scores_long <- melt(Map_Preditor_Hyper_Scores,
                                        variable.name = "Predictor",
                                        value.name    = "Score")

hypercomplementing <- ggplot(Map_Preditor_Hyper_Scores_long, aes(x = Predictor, y = Score)) +
  theme_classic() +
  geom_boxplot() +
  geom_jitter(color = alpha("black", 0.5), stroke = 0.5, width = 0.2) +
  ylab("Level of Functionality") +
  xlab("") +
  theme(legend.position = "none")

print(hypercomplementing)

ggsave(filename = "~/Desktop/hypercomplementing.pdf",
       plot = hypercomplementing, device = "pdf", width = 8, height = 8)
```


```{r, warning=FALSE, Dunham vs Baseline scores}
# =============================================================================
# CHUNK 9 — Dunham 33-Map Substitution Trend Comparison
#
# Tests whether the amino acid substitution trends in the CPOX baseline and
# metal maps are consistent with trends across 33 published DMS maps
# (Dunham & Beltrao, 2021).
#
# Focuses on the TO residue: for each destination amino acid, compares the
# median score across the CPOX maps and the Dunham aggregated matrix.
# Also compares baseline vs. metal directly.
# Spearman rank correlation is used throughout.
# =============================================================================

# ── Helper: pre-process missense variants ────────────────────────────────
process_map <- function(map) {
  map %>%
    filter(!str_detect(hgvs_pro, "=|Ter")) %>%
    mutate(
      from     = substr(hgvs_pro, 3, 5),
      pos      = str_extract(hgvs_pro, "\\d+"),
      changeTo = substr(hgvs_pro, nchar(hgvs_pro) - 2, nchar(hgvs_pro)),
      aa_type_From = case_when(
        from %in% c("Ala","Gly","Val","Leu","Ile","Met","Phe","Trp","Tyr") ~ "hydrophobic",
        from %in% c("Ser","Thr","Asn","Gln","Cys")                        ~ "polar",
        from %in% c("Asp","Glu","Arg","Lys","His")                        ~ "charged",
        from %in% c("Pro")                                                  ~ "proline",
        TRUE                                                                ~ "unknown"
      ),
      aa_type_To = case_when(
        changeTo %in% c("Ala","Gly","Val","Leu","Ile","Met","Phe","Trp","Tyr") ~ "hydrophobic",
        changeTo %in% c("Ser","Thr","Asn","Gln","Cys")                        ~ "polar",
        changeTo %in% c("Asp","Glu")                                           ~ "neg_charged",
        changeTo %in% c("Arg","Lys","His")                                     ~ "pos_charged",
        changeTo %in% c("Pro")                                                  ~ "proline",
        TRUE                                                                     ~ "unknown"
      )
    )
}

baseline_map       <- read_sheet("Baseline Map")
baseline_processed <- process_map(baseline_map)

# Median score per destination amino acid in baseline map
median_scores_baseline_to <- baseline_processed %>%
  group_by(changeTo) %>%
  summarise(median_score_baseline_to = median(score, na.rm = TRUE), .groups = "drop")

# Read Dunham matrix and set synonymous positions to NA
Dunham                 <- read_sheet("Dunham Maps")
AminoAcidChangesDunham <- Dunham[, 4:24]

AminoAcidChangesDunham <- AminoAcidChangesDunham %>%
  rowwise() %>%
  mutate(across(everything(), ~ ifelse(cur_column() == wt, NA, .))) %>%
  ungroup()

# Median per destination AA in Dunham matrix
median_Dunham <- AminoAcidChangesDunham %>%
  select(where(is.numeric)) %>%
  summarise(across(everything(), ~ median(.x, na.rm = TRUE))) %>%
  pivot_longer(everything(), names_to = "changeTo", values_to = "median_dunham_to")

# Convert single-letter AA codes to three-letter
aa_replacements <- setNames(
  c("Ala","Arg","Asn","Asp","Cys","Gln","Glu","Gly","His","Ile","Leu","Lys",
    "Met","Phe","Pro","Pyl","Ser","Thr","Trp","Tyr","Val"),
  c("A","R","N","D","C","Q","E","G","H","I","L","K","M","F","P","O","S","T","W","Y","V")
)
median_Dunham$changeTo <- str_replace_all(median_Dunham$changeTo, aa_replacements)
# Fix multi-substitution artefacts from sequential string replacement
median_Dunham$changeTo <- median_Dunham$changeTo %>%
  gsub("Glyln", "Gln", .) %>%
  gsub("Glylu", "Glu", .) %>%
  gsub("Prohe", "Phe", .)

aggregate_dunham_maps_toAAscores <- merge(median_Dunham, median_scores_baseline_to, by = "changeTo")

# Scatter plot with Spearman correlation annotation
compare_variables <- function(data, var1, var2, label_var, x_label, y_label) {
  cor_test <- cor.test(data[[var1]], data[[var2]], method = "spearman")
  p <- ggplot(data, aes(x = .data[[var1]], y = .data[[var2]])) +
    geom_point(alpha = 0.5) +
    theme_minimal() +
    labs(x = x_label, y = y_label) +
    ggtitle(paste("Correlation:", round(cor_test$estimate, 3),
                  "P-value:", format.pval(cor_test$p.value, digits = 3, format = "e"))) +
    geom_text_repel(aes(label = changeTo), size = 4, box.padding = 0.5)
  list(correlation = cor_test$estimate, p_value = cor_test$p.value)
  print(p)
}

result_baseline_Dunham <- compare_variables(
  aggregate_dunham_maps_toAAscores, "median_score_baseline_to", "median_dunham_to", "changeTo",
  "median score\n baseline map", "median score\n Dunham and Beltrao, 2021"
)

# ── Metal map comparisons ─────────────────────────────────────────────────
metal_map       <- read_sheet("Mercury Map")
metal_processed <- process_map(metal_map)

median_scores_metal_to <- metal_processed %>%
  group_by(changeTo) %>%
  summarise(median_score_metal_to = median(score, na.rm = TRUE), .groups = "drop")

aggregate_metal_maps_toAAscores <- merge(median_scores_baseline_to, median_scores_metal_to, by = "changeTo")
result_metal_baseline <- compare_variables(
  aggregate_metal_maps_toAAscores, "median_score_baseline_to", "median_score_metal_to", "changeTo",
  "median score\n baseline map", "median score\n metal map"
)

aggregate_dunham_metalmaps_toAAscores <- merge(median_Dunham, median_scores_metal_to, by = "changeTo")
result_metal_Dunham <- compare_variables(
  aggregate_dunham_metalmaps_toAAscores, "median_score_metal_to", "median_dunham_to", "changeTo",
  "median score\n metal map", "median score\n Dunham and Beltrao, 2021"
)
```


```{r, warning=FALSE, Impact Based on Substitution/AA Type}
# =============================================================================
# CHUNK 10 — Amino Acid Substitution Type Analysis (FROM / TO)
#
# Tests whether the physicochemical class of the source (FROM) or destination
# (TO) amino acid predicts variant functional impact.
#
# FROM analysis: excludes self-substitutions; uses broad categories
#   (hydrophobic / polar / charged / proline)
# TO analysis:   uses detailed categories (splits charged into +/- for TO)
#   and performs binary comparisons (e.g. "TO proline" vs. "not proline")
#
# Statistics: pairwise Wilcoxon with Bonferroni correction.
# Plots saved automatically to ~/Desktop/.
# =============================================================================

library(data.table); library(dplyr); library(stringr); library(ggplot2); library(ggpubr)

baseline <- read_sheet("Baseline Map")
metal    <- read_sheet("Mercury Map")

# Amino acid class definitions
aa_cats_basic <- list(
  hydrophobic = c("Ala","Gly","Val","Leu","Ile","Met","Phe","Trp","Tyr"),
  polar       = c("Ser","Thr","Asn","Gln","Cys"),
  charged     = c("Asp","Glu","Arg","Lys","His"),
  proline     = c("Pro")
)

# Detailed version: charged split into negative / positive for TO analysis
aa_cats_detailed <- list(
  hydrophobic = aa_cats_basic$hydrophobic,
  polar       = aa_cats_basic$polar,
  neg_charged = c("Asp","Glu"),
  pos_charged = c("Arg","Lys","His"),
  proline     = aa_cats_basic$proline
)

# Extract FROM/TO residues and assign amino acid class labels
process_map <- function(map, detailed = FALSE) {
  map_missense           <- map[!(grepl("=", map$hgvs_pro) | grepl("Ter", map$hgvs_pro)), ]
  map_missense$from      <- substr(map_missense$hgvs_pro, 3, 5)
  map_missense$pos       <- str_extract(map_missense$hgvs_pro, "[0-9]+")
  map_missense$changeTo  <- substr(map_missense$hgvs_pro,
                                    nchar(map_missense$hgvs_pro) - 2,
                                    nchar(map_missense$hgvs_pro))
  cats <- if (detailed) aa_cats_detailed else aa_cats_basic
  map_missense <- map_missense %>%
    mutate(
      aa_type_From = case_when(
        from %in% cats$hydrophobic ~ "hydrophobic",
        from %in% cats$polar ~ "polar",
        detailed & from %in% cats$neg_charged ~ "neg_charged",
        detailed & from %in% cats$pos_charged ~ "pos_charged",
        !detailed & from %in% cats$charged ~ "charged",
        from %in% cats$proline ~ "proline",
        TRUE ~ "unknown"
      ),
      aa_type_To = case_when(
        changeTo %in% cats$hydrophobic ~ "hydrophobic",
        changeTo %in% cats$polar ~ "polar",
        detailed & changeTo %in% cats$neg_charged ~ "neg_charged",
        detailed & changeTo %in% cats$pos_charged ~ "pos_charged",
        !detailed & changeTo %in% cats$charged ~ "charged",
        changeTo %in% cats$proline ~ "proline",
        TRUE ~ "unknown"
      )
    )
  return(map_missense)
}

# Unified analysis: pairwise Wilcoxon + boxplot with significance brackets
analyze_aa_category <- function(data, direction = c("FROM","TO"),
                                 simple_group = NULL, plot_title,
                                 pdf_filename, width = 6, height = 6) {
  direction <- match.arg(direction)
  aa_col    <- ifelse(direction == "FROM", "aa_type_From", "aa_type_To")
  
  if (direction == "FROM") data <- data[data$aa_type_From != data$aa_type_To, ]
  
  if (!is.null(simple_group)) {
    group_col <- paste0(aa_col, "_simple")
    data <- data %>%
      mutate(!!group_col := ifelse(get(aa_col) == simple_group, simple_group,
                                    paste0("not_", simple_group)))
    data[[group_col]] <- factor(data[[group_col]],
                                levels = c(simple_group, paste0("not_", simple_group)))
  } else {
    group_col         <- aa_col
    data[[group_col]] <- factor(data[[group_col]])
  }
  
  categories <- levels(data[[group_col]])
  if (length(categories) < 2) return(NULL)
  
  results <- data.frame()
  for (i in 1:(length(categories) - 1)) {
    for (j in (i + 1):length(categories)) {
      cat1 <- categories[i]; cat2 <- categories[j]
      g1   <- data$score[data[[group_col]] == cat1]
      g2   <- data$score[data[[group_col]] == cat2]
      test <- wilcox.test(g1, g2, correct = TRUE)
      results <- rbind(results, data.frame(
        Category1 = cat1, Category2 = cat2,
        p.value   = test$p.value,
        Median1   = median(g1, na.rm = TRUE),
        Median2   = median(g2, na.rm = TRUE)
      ))
    }
  }
  results$p.adj <- p.adjust(results$p.value, method = "bonferroni")
  
  p <- ggplot(data, aes_string(x = group_col, y = "score")) +
    geom_boxplot(outlier.size = 0.3) +
    labs(title = plot_title, x = paste(direction, "amino acid category"),
         y = "Variant functional score") +
    theme_minimal(base_size = 17)
  
  sig_results <- results[results$p.adj < 0.05, ]
  if (nrow(sig_results) > 0) {
    comps <- mapply(c, sig_results$Category1, sig_results$Category2, SIMPLIFY = FALSE)
    p <- p + stat_compare_means(comparisons = comps, method = "wilcox.test",
                                label = "p.signif", p.adjust.method = "bonferroni")
  }
  
  ggsave(paste0("~/Desktop/", pdf_filename), plot = p, device = "pdf",
         width = width, height = height)
  print(p)
}

baseline_basic    <- process_map(baseline, FALSE)
metal_basic       <- process_map(metal,    FALSE)
baseline_detailed <- process_map(baseline, TRUE)
metal_detailed    <- process_map(metal,    TRUE)

# FROM analyses (broad categories)
analyze_aa_category(baseline_basic, "FROM", NULL, "Baseline map: FROM amino acid type", "Baseline_FROM_full.pdf")
analyze_aa_category(metal_basic,    "FROM", NULL, "Metal map: FROM amino acid type",    "Metal_FROM_full.pdf")

# TO analyses — baseline (binary comparisons per class)
analyze_aa_category(baseline_detailed, "TO", "polar",       "Baseline map: TO polar vs others",           "Baseline_TO_polar.pdf")
analyze_aa_category(baseline_detailed, "TO", "hydrophobic", "Baseline map: TO hydrophobic vs others",     "Baseline_TO_hydrophobic.pdf")
analyze_aa_category(baseline_detailed, "TO", "proline",     "Baseline map: TO proline vs others",         "Baseline_TO_proline.pdf")
analyze_aa_category(baseline_detailed, "TO", "neg_charged", "Baseline map: TO negative charged vs others","Baseline_TO_neg_charged.pdf")

# TO analyses — metal map
analyze_aa_category(metal_detailed, "TO", "polar",       "Metal map: TO polar vs others",           "Metal_TO_polar.pdf")
analyze_aa_category(metal_detailed, "TO", "hydrophobic", "Metal map: TO hydrophobic vs others",     "Metal_TO_hydrophobic.pdf")
analyze_aa_category(metal_detailed, "TO", "proline",     "Metal map: TO proline vs others",         "Metal_TO_proline.pdf")
analyze_aa_category(metal_detailed, "TO", "neg_charged", "Metal map: TO negative charged vs others","Metal_TO_neg_charged.pdf")
```


```{r, warning=FALSE, Impact Based on Quaternary Structure}
# =============================================================================
# CHUNK 11 — Structure-Based Analysis (SASA / Quaternary Domains)
#
# Tests whether positional median DMS scores differ between structural contexts:
#   • core          : buried (SASA < 25 Å²), excluding functional sites
#   • surface       : exposed (SASA > 50 Å²), excluding functional sites
#   • active site   : catalytic residues
#   • dimer interface: residues at the CPOX homodimer interface
#   • MTS           : mitochondrial targeting sequence (residues 1–110)
#
# Pairwise Wilcoxon tests are performed for both baseline and metal maps.
# Results are visualised as boxplots and a triangular heatmap (baseline =
# upper-left triangle, metal = lower-right triangle).
# =============================================================================

library(data.table); library(dplyr); library(stringr); library(tidyr)
library(ggplot2); library(ggpubr)

# SASA sheet: row 1 = description, row 2 = column headers (read_sheet handles this).
# Columns: AA residue | AA position | Total SASA (Å²) | Total relative SASA (%) |
#          Main-chain SASA (Å²) | Main-chain relative SASA (%) |
#          Side-chain SASA (Å²) | Side-chain relative SASA (%) |
#          Polar SASA (Å²) | Polar relative SASA (%) |
#          Apolar SASA (Å²) | Apolar relative SASA (%)
freesasa_data <- read_sheet("SASA") %>%
  transmute(
    from_aa = suppressWarnings(as.numeric(`AA position`)),
    sasa    = suppressWarnings(as.numeric(`Total SASA (Å²)`))
  ) %>%
  filter(!is.na(from_aa), !is.na(sasa))

# Load position lists for excluded functional regions
ignore_positions <- list(
  AS  = read_sheet("Active Site Residues")$'AA position',
  DI  = read_sheet("Interface Residues")$'AA position',
  MTS = 1:110
)

# Core: buried residues not in any functional region
core_pos <- freesasa_data %>%
  filter(sasa < 25, !from_aa %in% unlist(ignore_positions)) %>%
  pull(from_aa)

# Surface: exposed residues not in any functional region
surface_pos <- freesasa_data %>%
  filter(sasa > 50, !from_aa %in% unlist(ignore_positions)) %>%
  pull(from_aa)

# Load map and annotate each position with its structural context
load_map <- function(file, label) {
  sheet <- c(baseline = "Baseline Map", metal = "Mercury Map")[label]
  read_sheet(sheet) %>%
    mutate(pos = as.numeric(str_extract(hgvs_pro, "\\d+"))) %>%
    group_by(pos) %>%
    summarise(median_score = median(score, na.rm = TRUE), .groups = "drop") %>%
    mutate(
      residue_type = case_when(
        pos %in% surface_pos          ~ "surface",
        pos %in% core_pos             ~ "core",
        pos %in% ignore_positions$DI  ~ "dimer interface",
        pos %in% ignore_positions$MTS ~ "MTS",
        pos %in% ignore_positions$AS  ~ "active site",
        TRUE                          ~ NA_character_
      ),
      map_label = label
    ) %>%
    filter(!is.na(residue_type))
}

baseline_res <- load_map("~/Desktop/Manuscript_Data.xlsx", "baseline")
metal_res    <- load_map("~/Desktop/Manuscript_Data.xlsx",   "metal")

combined_res  <- bind_rows(baseline_res, metal_res) %>%
  mutate(map_type = tolower(map_label))

residue_levels <- c("active site","core","dimer interface","MTS","surface")
combined_res$residue_type <- factor(combined_res$residue_type, levels = residue_levels)

# All pairwise group combinations for Wilcoxon tests
all_combinations <- expand.grid(
  group1 = unique(combined_res$residue_type),
  group2 = unique(combined_res$residue_type),
  stringsAsFactors = FALSE
)

# Perform Wilcoxon for both maps across all pairs
perform_wilcoxon_test <- function(data, combos) {
  nested <- data %>%
    group_by(map_label, residue_type) %>%
    summarise(scores = list(median_score), med = median(median_score, na.rm = TRUE),
              .groups = "drop") %>%
    pivot_wider(names_from = map_label, values_from = c(scores, med), names_sep = "_")
  
  combos %>%
    rowwise() %>%
    mutate(
      p_value_baseline = wilcox.test(
        nested %>% filter(residue_type == group1) %>% pull(scores_baseline) %>% .[[1]],
        nested %>% filter(residue_type == group2) %>% pull(scores_baseline) %>% .[[1]],
        exact = FALSE)$p.value,
      p_value_metal = wilcox.test(
        nested %>% filter(residue_type == group1) %>% pull(scores_metal) %>% .[[1]],
        nested %>% filter(residue_type == group2) %>% pull(scores_metal) %>% .[[1]],
        exact = FALSE)$p.value,
      diff_med_baseline = (nested %>% filter(residue_type == group1) %>% pull(med_baseline)) -
                          (nested %>% filter(residue_type == group2) %>% pull(med_baseline)),
      diff_med_metal    = (nested %>% filter(residue_type == group1) %>% pull(med_metal)) -
                          (nested %>% filter(residue_type == group2) %>% pull(med_metal))
    ) %>% ungroup()
}

wilcoxon_results <- perform_wilcoxon_test(combined_res, all_combinations)

# Convert p-values to ordered significance levels (0–3)
annotate_significance <- function(df, p_col, map_label, diff_col) {
  df %>%
    filter(!is.na(.data[[p_col]])) %>%
    mutate(
      significance = case_when(
        .data[[p_col]] < 0.001 ~ "p < 0.001",
        .data[[p_col]] < 0.01  ~ "p < 0.01",
        .data[[p_col]] < 0.05  ~ "p < 0.05",
        TRUE ~ "ns"
      ),
      significance_num = case_when(
        .data[[p_col]] < 0.001 ~ 3,
        .data[[p_col]] < 0.01  ~ 2,
        .data[[p_col]] < 0.05  ~ 1,
        TRUE ~ 0
      ),
      map_type    = map_label,
      diff_median = .data[[diff_col]]
    ) %>%
    select(group1, group2, map_type, p_value = all_of(p_col), significance, significance_num, diff_median)
}

wilcoxon_results_combined <- bind_rows(
  annotate_significance(wilcoxon_results, "p_value_baseline", "baseline", "diff_med_baseline"),
  annotate_significance(wilcoxon_results, "p_value_metal",    "metal",    "diff_med_metal")
)

# SASA histogram with core/surface cut-off lines
p_hist <- ggplot(freesasa_data, aes(x = sasa)) +
  geom_histogram(binwidth = 5, fill = "grey", color = "black") +
  geom_vline(xintercept = c(25, 50), color = c("magenta","green"), linetype = "dashed") +
  labs(title = "freeSASA", x = "SASA (Å²)", y = "Frequency") +
  theme_minimal(base_size = 15)
print(p_hist)
pdf("~/Desktop/freeSASA_histogram.pdf", width = 7, height = 6); print(p_hist); dev.off()

# Significance comparisons to annotate on boxplot
sig_comps <- list(
  c("surface", "core"),
  c("surface", "active site"),
  c("core", "dimer interface")
)

ylim_max <- max(combined_res$median_score, na.rm = TRUE) + 0.75

# Combined boxplot (baseline and metal side by side per structural category)
p_combined <- ggplot(combined_res, aes(x = residue_type, y = median_score, fill = map_type)) +
  geom_boxplot(outlier.shape = NA, position = position_dodge(width = 0.75)) +
  scale_fill_manual(values = c("baseline" = "#11EE89", "metal" = "#EE1176")) +
  stat_compare_means(comparisons = sig_comps, method = "wilcox.test", label = "p.signif") +
  coord_cartesian(ylim = c(0, ylim_max)) +
  labs(x = "Residue Type", y = "Functional Score (Median)", fill = "Map Type") +
  theme_minimal(base_size = 15)
print(p_combined)
pdf("~/Desktop/Combined_Baseline_Metal_Boxplot.pdf", width = 8, height = 6); print(p_combined); dev.off()

# Separate boxplots per map
for (map in c("baseline", "metal")) {
  df_map     <- combined_res %>% filter(map_label == map)
  fill_color <- ifelse(map == "baseline", "#11EE89", "#EE1176")
  p <- ggplot(df_map, aes(x = residue_type, y = median_score)) +
    geom_boxplot(outlier.shape = NA, fill = fill_color) +
    scale_x_discrete(drop = FALSE, limits = residue_levels) +
    stat_compare_means(comparisons = sig_comps, method = "wilcox.test", label = "p.signif") +
    coord_cartesian(ylim = c(0, ylim_max)) +
    labs(title = paste0(str_to_title(map), " Map"),
         x = "Residue Type", y = "Functional Score (Median)") +
    theme_minimal(base_size = 15)
  pdf(paste0("~/Desktop/", str_to_title(map), "_Map_Boxplot.pdf"), width = 8, height = 6)
  print(p); dev.off()
}

# ── Triangular heatmap ────────────────────────────────────────────────────
# Each cell is split diagonally: upper-left = baseline, lower-right = metal.
# Fill encodes significance level (white → light pink → red → dark red).
heatmap_matrix <- wilcoxon_results_combined %>%
  select(group1, group2, map_type, significance_num) %>%
  pivot_wider(names_from = map_type, values_from = significance_num, values_fill = 0) %>%
  rename(residue1 = group1, residue2 = group2) %>%
  mutate(x = factor(residue1, levels = residue_levels),
         y = factor(residue2, levels = residue_levels))

heatmap_matrix$x_num <- as.numeric(heatmap_matrix$x)
heatmap_matrix$y_num <- as.numeric(heatmap_matrix$y)

# Build triangle polygon coordinates for each half of each cell
make_triangles <- function(data, type = c("baseline","metal")) {
  type   <- match.arg(type)
  offset <- if (type == "baseline") list(x = c(-0.5, 0.5,-0.5), y = c( 0.5, 0.5,-0.5))
            else                    list(x = c( 0.5,-0.5, 0.5), y = c(-0.5,-0.5, 0.5))
  data[rep(1:nrow(data), each = 3), ] %>%
    mutate(
      x    = rep(data$x_num, each = 3) + rep(offset$x, times = nrow(data)),
      y    = rep(data$y_num, each = 3) + rep(offset$y, times = nrow(data)),
      fill = rep(data[[type]], each = 3)
    ) %>%
    select(x, y, fill, residue1, residue2)
}

baseline_triangles <- make_triangles(heatmap_matrix, "baseline")
metal_triangles    <- make_triangles(heatmap_matrix, "metal")

p_heatmap <- ggplot() +
  geom_polygon(data = metal_triangles,
               aes(x = x, y = y, group = interaction(residue1, residue2), fill = factor(fill)),
               color = "black") +
  geom_polygon(data = baseline_triangles,
               aes(x = x, y = y, group = interaction(residue1, residue2), fill = factor(fill)),
               color = "black") +
  coord_fixed() +
  scale_fill_manual(
    values = c("0"="white","1"="#FFDDDD","2"="#FF8888","3"="#B20000"),
    breaks = c("0","1","2","3"),
    labels = c("Not Sign.","< 0.05","< 0.01","< 0.001"),
    name   = "Significance (p.value)"
  ) +
  scale_x_continuous(breaks = seq_along(residue_levels), labels = residue_levels,
                     expand = expansion(add = 0.5), position = "top") +
  scale_y_continuous(breaks = seq_along(residue_levels), labels = residue_levels,
                     expand = expansion(add = 0.5), trans = "reverse") +
  labs(title = "Triangular Heatmap for Metal and Baseline") +
  theme_minimal(base_size = 14) +
  theme(axis.title = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 0),
        axis.text.y = element_text(hjust = 1),
        panel.grid = element_blank())
print(p_heatmap)
pdf("~/Desktop/Triangular_Heatmap_Baseline_Metal.pdf", width = 8, height = 8)
print(p_heatmap); dev.off()
```


```{r, warning=FALSE, Baseline and Metal: Impact Based on Quaternary Position: MTS}
# =============================================================================
# CHUNK 12 — MTS Region Structure Analysis
#
# Focuses specifically on the mitochondrial targeting sequence (MTS, residues
# 1–110) and its sub-features:
#   • TOM20 motif   : residues 12, 14, 15 — receptor binding
#   • Amphipathic helix: 75–93
#   • MPP cleavage motif: 36–40 (matrix processing peptidase recognition)
#   • Mature N-termini:  109–111
#   • Surface / dimer interface from SASA (excluded from MTS-specific analysis)
#
# Tests whether these MTS sub-features show distinct functional score
# distributions compared to surface positions.
# =============================================================================

# SASA sheet: row 1 = description, row 2 = column headers (read_sheet handles this).
freesasa_data <- read_sheet("SASA") %>%
  transmute(
    from_aa = suppressWarnings(as.numeric(`AA position`)),
    sasa    = suppressWarnings(as.numeric(`Total SASA (Å²)`))
  ) %>%
  filter(!is.na(from_aa), !is.na(sasa))

ignore_positions <- list(
  AS  = read_sheet("Active Site Residues")$'AA position',  # was $AA — column is "AA position"
  DI  = read_sheet("Interface Residues")$'AA position',    # was $AA — column is "AA position"
  MTS = 1:110
)

surface_positions <- freesasa_data %>%
  filter(sasa > 50, !from_aa %in% unlist(ignore_positions)) %>%
  pull(from_aa)

# Load map and extract position, WT amino acid, and median score per position
load_and_process_map <- function(file, sheet, map_type) {
  read_sheet(sheet) %>%
    mutate(
      pos        = as.numeric(gsub("p\\.([A-Za-z]+)([0-9]+)([A-Za-z]+)", "\\2", hgvs_pro)),
      mutated_aa = gsub("p\\.([A-Za-z]+)([0-9]+)([A-Za-z]+)", "\\1", hgvs_pro),
      map_type   = map_type
    ) %>%
    group_by(pos) %>%
    mutate(median_score = median(score, na.rm = TRUE)) %>%
    select(pos, mutated_aa, map_type, median_score)
}

baseline_data <- load_and_process_map("~/Desktop/Manuscript_Data.xlsx", "Baseline Map", "baseline")
metal_data    <- load_and_process_map("~/Desktop/Manuscript_Data.xlsx",   "Mercury Map",  "metal")

combined_data <- bind_rows(baseline_data, metal_data) %>%
  distinct() %>%
  filter(!is.na(pos) & !is.na(mutated_aa) & !is.na(median_score))

# Define MTS sub-feature position sets
TOM20motif_positions  <- c(12, 14, 15)
amphipathicHelix      <- 75:93
Ntermini_positions    <- 109:111
MPP_positions         <- 36:40

# Annotate positions with MTS sub-feature labels
residue_data <- combined_data %>%
  mutate(
    residue_type = case_when(
      pos %in% surface_positions        ~ "surface",
      pos %in% ignore_positions$DI      ~ "dimer interface",
      pos %in% TOM20motif_positions     ~ "TOM20",
      pos %in% MPP_positions            ~ "MPP motif",
      pos %in% amphipathicHelix         ~ "helix",
      pos %in% Ntermini_positions       ~ "N-termini",
      TRUE                              ~ NA_character_
    )
  ) %>%
  filter(!is.na(residue_type)) %>%
  mutate(residue_type = factor(residue_type,
                               levels = c("surface","dimer interface","TOM20","MPP motif","helix","N-termini")))

MTS_boxplot <- ggplot(residue_data, aes(x = residue_type, y = median_score, fill = map_type)) +
  geom_boxplot(outlier.shape = NA, position = position_dodge(width = 0.75)) +
  scale_x_discrete(labels = c(
    "surface"         = "surface",
    "dimer interface" = "dimer interface",
    "TOM20"           = "TOM20",
    "MPP motif"       = "MPP motif",
    "helix"           = "α-helix",
    "N-termini"       = "Mature \nN-termini"
  )) +
  scale_fill_manual(values = c("baseline" = "#11EE89", "metal" = "#EE1176")) +
  labs(x = "residue type/pos", y = "functional score (median)") +
  theme_minimal()

pdf("~/Desktop/MTS_boxplots.pdf", width = 8, height = 8); print(MTS_boxplot); dev.off()
print(MTS_boxplot)

# Pairwise Wilcoxon tests for MTS sub-features
all_combinations <- expand.grid(
  group1 = unique(residue_data$residue_type),
  group2 = unique(residue_data$residue_type),
  stringsAsFactors = FALSE
)

perform_wilcoxon_test <- function(residue_data, all_combinations) {
  nested_data <- residue_data %>%
    group_by(map_type, residue_type) %>%
    summarise(scores = list(median_score), .groups = "drop") %>%
    pivot_wider(names_from = map_type, values_from = scores)
  
  wilcoxon_if_possible <- function(x, y) {
    if (length(x) >= 2 && length(y) >= 2) wilcox.test(x, y, exact = FALSE)$p.value else NA
  }
  
  all_combinations %>%
    rowwise() %>%
    mutate(
      p_value_baseline = {
        x <- nested_data %>% filter(residue_type == group1) %>% pull(baseline) %>% .[[1]]
        y <- nested_data %>% filter(residue_type == group2) %>% pull(baseline) %>% .[[1]]
        wilcoxon_if_possible(x %||% numeric(0), y %||% numeric(0))
      },
      p_value_metal = {
        x <- nested_data %>% filter(residue_type == group1) %>% pull(metal) %>% .[[1]]
        y <- nested_data %>% filter(residue_type == group2) %>% pull(metal) %>% .[[1]]
        wilcoxon_if_possible(x %||% numeric(0), y %||% numeric(0))
      }
    ) %>% ungroup()
}

wilcoxon_results <- perform_wilcoxon_test(residue_data, all_combinations)

annotate_significance <- function(df, p_col, map_label) {
  df %>%
    filter(!is.na(.data[[p_col]])) %>%
    mutate(
      significance = case_when(
        .data[[p_col]] < 0.001 ~ "p < 0.001",
        .data[[p_col]] < 0.01  ~ "p < 0.01",
        .data[[p_col]] < 0.05  ~ "p < 0.05",
        TRUE ~ "ns"
      ),
      significance_num = case_when(
        .data[[p_col]] < 0.001 ~ 3,
        .data[[p_col]] < 0.01  ~ 2,
        .data[[p_col]] < 0.05  ~ 1,
        TRUE ~ 0
      ),
      map_type = map_label
    ) %>%
    select(group1, group2, map_type, p_value = all_of(p_col), significance, significance_num)
}

wilcoxon_results_combined <- bind_rows(
  annotate_significance(wilcoxon_results, "p_value_baseline", "baseline"),
  annotate_significance(wilcoxon_results, "p_value_metal",    "metal")
)

# Reshape for triangular heatmap
heatmap_matrix <- wilcoxon_results_combined %>%
  select(group1, group2, map_type, significance_num) %>%
  pivot_wider(names_from = map_type, values_from = significance_num, values_fill = 0) %>%
  rename(residue1 = group1, residue2 = group2) %>%
  mutate(x = factor(residue1), y = factor(residue2))

heatmap_matrix$x_num <- as.numeric(heatmap_matrix$x)
heatmap_matrix$y_num <- as.numeric(heatmap_matrix$y)

make_triangles <- function(data, type = c("baseline","metal")) {
  type   <- match.arg(type)
  offset <- if (type == "baseline") list(x = c(-0.5, 0.5,-0.5), y = c( 0.5, 0.5,-0.5))
            else                    list(x = c( 0.5,-0.5, 0.5), y = c(-0.5,-0.5, 0.5))
  data[rep(1:nrow(data), each = 3), ] %>%
    mutate(
      x    = rep(data$x_num, each = 3) + rep(offset$x, times = nrow(data)),
      y    = rep(data$y_num, each = 3) + rep(offset$y, times = nrow(data)),
      fill = rep(data[[type]], each = 3)
    ) %>%
    select(x, y, fill, residue1, residue2)
}

baseline_triangles <- make_triangles(heatmap_matrix, "baseline")
metal_triangles    <- make_triangles(heatmap_matrix, "metal")
baseline_triangles$fill <- as.numeric(baseline_triangles$fill)
metal_triangles$fill    <- as.numeric(metal_triangles$fill)

residue_levels <- factor(
  unique(c(as.character(baseline_triangles$residue1), as.character(metal_triangles$residue1))),
  levels = c("surface","dimer interface","TOM20","MPP motif","helix","N-termini")
)

baseline_triangles <- baseline_triangles %>%
  mutate(residue1 = factor(residue1, levels = residue_levels),
         residue2 = factor(residue2, levels = residue_levels))
metal_triangles <- metal_triangles %>%
  mutate(residue1 = factor(residue1, levels = residue_levels),
         residue2 = factor(residue2, levels = residue_levels))

MTS_Heatmap <- ggplot() +
  geom_polygon(data = metal_triangles,
               aes(x = x, y = y, group = interaction(residue1, residue2), fill = factor(fill)),
               color = "black", size = 0.5) +
  geom_polygon(data = baseline_triangles,
               aes(x = x, y = y, group = interaction(residue1, residue2), fill = factor(fill)),
               color = "black", size = 0.5) +
  coord_fixed() +
  scale_fill_manual(
    values = c("0"="white","1"="#FFDDDD","2"="#FF8888","3"="#B20000"),
    breaks = c("0","1","2","3"),
    labels = c("Not Sign.","< 0.05","< 0.01","< 0.001"),
    name   = "Significance (p.value)"
  ) +
  scale_x_discrete(limits = levels(residue_levels), labels = levels(residue_levels)) +
  scale_y_discrete(limits = levels(residue_levels), labels = levels(residue_levels)) +
  geom_text(data = data.frame(x = levels(residue_levels),
                               y = rep(-2, length(levels(residue_levels)))),
            aes(x = x, y = y, label = x), size = 4, color = "black", angle = 90, hjust = 0) +
  geom_text(data = data.frame(x = rep(-2, length(levels(residue_levels))),
                               y = levels(residue_levels)),
            aes(x = x, y = y, label = y), size = 4, color = "black", angle = 0, hjust = 0) +
  theme_minimal() +
  theme(axis.text.x = element_blank(), axis.text.y = element_blank(),
        panel.grid = element_blank(),
        legend.key = element_rect(fill = "white", color = "black")) +
  labs(title = "Triangular Heatmap for Metal and Baseline",
       x = "Residue Type", y = "Residue Type", fill = "Significance")

pdf("~/Desktop/Heatmap_MTS.pdf", width = 8, height = 8); print(MTS_Heatmap); dev.off()
print(MTS_Heatmap)

# Print pairwise comparisons of specific groups for both maps
print_comparison <- function(data, map_label, group1, group2) {
  scores1 <- data %>% filter(map_type == map_label, residue_type == group1) %>% pull(median_score)
  scores2 <- data %>% filter(map_type == map_label, residue_type == group2) %>% pull(median_score)
  cat(sprintf("\n%s map - %s group size: %d\n", map_label, group1, length(scores1)))
  cat(sprintf("%s map - %s group size: %d\n", map_label, group2, length(scores2)))
  median_diff <- median(scores1, na.rm = TRUE) - median(scores2, na.rm = TRUE)
  wilcox_res  <- wilcox.test(scores1, scores2, exact = FALSE)
  cat(sprintf("%s map comparison: %s vs %s\n", toupper(map_label), group1, group2))
  cat(sprintf("Difference in median functional scores: %.4f\n", median_diff))
  cat(sprintf("Wilcoxon test p-value: %.4g\n", wilcox_res$p.value))
}

print_comparison(residue_data, "baseline", "helix", "surface")
print_comparison(residue_data, "metal",    "helix", "surface")
```


```{r, warning=FALSE, Substitution Resulting in Negative Charge for Second Decarboxylation Residues}
# =============================================================================
# CHUNK 13 — Negative-Charge Substitutions at Second Decarboxylation Residues
#
# CPOX catalyses two sequential decarboxylation reactions. Residues involved
# in each reaction occupy distinct spatial positions:
#   set1 (400–404): second decarboxylation substrate-binding residues
#   set2 (407, 411, 413, 416): second set
#
# This chunk tests whether substitutions TO negatively charged amino acids
# (Asp/Glu) at these specific residue sets are significantly more disruptive
# than:
#   1) All other substitutions at other positions (regardless of AA type)
#   2) TO negative-charge substitutions at all other positions
#
# One-tailed Wilcoxon (alternative = "less") tests whether target scores
# are lower than the comparison group.
# =============================================================================

library(data.table); library(stringr); library(ggplot2); library(ggpubr)

baseline <- read_sheet("Baseline Map")
setDT(baseline)
metal    <- read_sheet("Mercury Map")
setDT(metal)

aa_cats_detailed <- list(
  hydrophobic = c("Ala","Gly","Val","Leu","Ile","Met","Phe","Trp","Tyr"),
  polar       = c("Ser","Thr","Asn","Gln","Cys"),
  neg_charged = c("Asp","Glu"),
  pos_charged = c("Arg","Lys","His"),
  proline     = c("Pro")
)

# Lookup table: map each AA to its physicochemical class
aa_lookup <- rbindlist(
  lapply(names(aa_cats_detailed), function(cat) {
    data.table(changeTo = aa_cats_detailed[[cat]], aa_type_To_simple = cat)
  })
)

set1 <- 400:404      # first decarboxylation residue set
set2 <- c(407, 411, 413, 416)  # second set

# Main analysis function: two comparisons per dataset × target set
#   Comparison 1: neg-charged @ target vs. all other substitutions
#   Comparison 2: neg-charged @ target vs. neg-charged at all other positions
process_neg_target_vs_all <- function(df, target_set, exclude_set,
                                       dataset_name = "dataset",
                                       pdf_file_prefix = NULL) {
  setDT(df)
  df[, pos      := as.numeric(str_extract(hgvs_pro, "\\d+"))]
  df[, changeTo := str_extract(hgvs_pro, "(?<=\\d)[A-Z][a-z]{2}$")]
  df <- merge(df, aa_lookup, by = "changeTo", all.x = TRUE)
  
  # Comparison 1: target neg-charged vs. all other (any type, any position)
  neg_target_all <- df[pos %in% target_set & aa_type_To_simple == "neg_charged"]
  all_others_all <- df[!(pos %in% target_set & aa_type_To_simple == "neg_charged") &
                         !(pos %in% exclude_set & aa_type_To_simple == "neg_charged")]
  
  data_all <- rbind(copy(neg_target_all)[, group := "neg_target"],
                    copy(all_others_all)[, group := "all_other"])
  data_all[, group := factor(group, levels = c("neg_target","all_other"))]
  
  medians_all    <- data_all[, .(median_score = median(score, na.rm = TRUE)), by = group]
  median_diff_all <- medians_all[group == "neg_target", median_score] -
                     medians_all[group == "all_other",  median_score]
  test_all <- wilcox.test(score ~ group, data = data_all, alternative = "less", exact = FALSE)
  
  if (!is.null(pdf_file_prefix)) {
    p_all <- ggplot(data_all, aes(x = group, y = score)) +
      geom_boxplot(outlier.size = 0.3) +
      labs(title = paste0(dataset_name, ": neg-charged @ ",
                          paste(target_set, collapse = ","), " vs all other residues"),
           x = "Group", y = "Variant functional score") +
      theme_minimal(base_size = 16) +
      stat_compare_means(method = "wilcox.test", label = "p.signif")
    ggsave(paste0(pdf_file_prefix, "_all_substitutions.pdf"), p_all, width = 6, height = 6)
  }
  
  # Comparison 2: target neg-charged vs. neg-charged at all other positions
  neg_target_neg <- df[pos %in% target_set & aa_type_To_simple == "neg_charged"]
  neg_other_neg  <- df[pos %notin% target_set & aa_type_To_simple == "neg_charged"]
  
  data_neg <- rbind(copy(neg_target_neg)[, group := "neg_target"],
                    copy(neg_other_neg)[,  group := "neg_other"])
  data_neg[, group := factor(group, levels = c("neg_target","neg_other"))]
  
  medians_neg    <- data_neg[, .(median_score = median(score, na.rm = TRUE)), by = group]
  median_diff_neg <- medians_neg[group == "neg_target", median_score] -
                     medians_neg[group == "neg_other",  median_score]
  test_neg <- wilcox.test(score ~ group, data = data_neg, alternative = "less", exact = FALSE)
  
  if (!is.null(pdf_file_prefix)) {
    p_neg <- ggplot(data_neg, aes(x = group, y = score)) +
      geom_boxplot(outlier.size = 0.3) +
      labs(title = paste0(dataset_name, ": neg-charged @ ",
                          paste(target_set, collapse = ","), " vs all other TO negative"),
           x = "Group", y = "Variant functional score") +
      theme_minimal(base_size = 16) +
      stat_compare_means(method = "wilcox.test", label = "p.signif")
    ggsave(paste0(pdf_file_prefix, "_neg_only.pdf"), p_neg, width = 6, height = 6)
  }
  
  list(dataset = dataset_name, target = paste(target_set, collapse = ","),
       median_diff_all = median_diff_all, p_value_all = test_all$p.value,
       median_diff_neg = median_diff_neg, p_value_neg = test_neg$p.value)
}

res1 <- process_neg_target_vs_all(baseline, set1, set2, "baseline_set1", "~/Desktop/baseline_set1")
res2 <- process_neg_target_vs_all(baseline, set2, set1, "baseline_set2", "~/Desktop/baseline_set2")
res3 <- process_neg_target_vs_all(metal,    set1, set2, "metal_set1",    "~/Desktop/metal_set1")
res4 <- process_neg_target_vs_all(metal,    set2, set1, "metal_set2",    "~/Desktop/metal_set2")

for (r in list(res1, res2, res3, res4)) {
  cat("\nDataset:", r$dataset,
      "\nTarget positions:", r$target,
      "\nMedian difference (vs all other substitutions):", round(r$median_diff_all, 4),
      "\nOne-tailed Wilcoxon p (vs all other substitutions):", signif(r$p_value_all, 4),
      "\nMedian difference (vs all other TO negative):", round(r$median_diff_neg, 4),
      "\nOne-tailed Wilcoxon p (vs all other TO negative):", signif(r$p_value_neg, 4),
      "\n----------------------------------------------\n")
}
```


```{r, warning=FALSE, Impact of AA substitution and Substrate Preference with Hg}
# =============================================================================
# CHUNK 14 — Delta Score Impact Classification (APR / Improved / Known)
#
# Classifies variants by their response to mercury (Hg) exposure using
# delta scores (baseline – metal):
#
#   APR (activity reduced in presence of Hg):
#     • Significant delta (p < threshold)
#     • Negative deltaScore (function lost under metal)
#     • Tolerated at baseline (lower CI ≥ 0.5)
#
#   Improved (Hg-rescued):
#     • Significant delta
#     • Positive deltaScore (function gained under metal)
#     • Tolerated at baseline
#
#   Known (no significant delta):
#     • Not significant but tolerated at baseline
#
#   Neutral: all other variants
#
# Produces violin + boxplot distributions with APR variants highlighted,
# relative to a null band derived from synonymous + nonsense delta scores.
# =============================================================================

delta <- read_sheet("Delta Map")
setDT(delta)
delta <- delta[is.finite(deltaScore) & is.finite(deltaSE)]

# Classify variant types
delta[, variant_type  := "Missense"]
delta[grepl("Ter$", hgvs_pro), variant_type := "Nonsense"]
delta[grepl("=$",   hgvs_pro), variant_type := "Synonymous"]
delta[, variant_group := variant_type]
delta[variant_type %in% c("Synonymous","Nonsense"), variant_group := "Synonymous + Nonsense"]

# Load baseline scores and define tolerance threshold
baseline_map <- read_sheet("Baseline Map")
setDT(baseline_map)
setnames(baseline_map, colnames(baseline_map)[1:4], c("hgvs_pro","baseline_score","se","df"))
baseline_map[, baseline_lower := baseline_score + se]

baseline_miss          <- !grepl("Ter$|=$", baseline_map$hgvs_pro)
baseline_miss_tolerated <- baseline_map[baseline_miss & baseline_lower >= 0.5, hgvs_pro]

# Assign impact categories
delta[, Impact := "neutral"]
APR_set      <- delta[significant == TRUE  & deltaScore < 0 & hgvs_pro %in% baseline_miss_tolerated, hgvs_pro]
Improved_set <- delta[significant == TRUE  & deltaScore > 0 & hgvs_pro %in% baseline_miss_tolerated, hgvs_pro]
Known_set    <- delta[significant == FALSE & hgvs_pro %in% baseline_miss_tolerated, hgvs_pro]

delta[hgvs_pro %in% APR_set,      Impact := "APR"]
delta[hgvs_pro %in% Improved_set, Impact := "Improved"]
delta[hgvs_pro %in% Known_set,    Impact := "Known"]

# Export impact classification for PyMOL structural colouring
color <- delta[, .(hgvs_pro, Impact)]
color[, score := NA_real_]
color[Impact == "Improved", score := 2]
color[Impact == "neutral",  score := 1.1]
color[Impact == "APR",      score := 0.6]
color[Impact == "Known",    score := 0]
color[, se  := 0]
color[, pos := as.numeric(str_extract(hgvs_pro, "[0-9]+"))]
write.csv(color[, .(hgvs_pro, score, se)], "~/Desktop/CPOX_delta_structure.csv", row.names = FALSE)

# Null distribution: synonymous + nonsense delta scores
null_variants <- delta[variant_group == "Synonymous + Nonsense"]
null_median   <- median(null_variants$deltaScore, na.rm = TRUE)
null_se       <- sd(null_variants$deltaScore, na.rm = TRUE) / sqrt(nrow(null_variants))
null_lower    <- null_median - 1.96 * null_se
null_upper    <- null_median + 1.96 * null_se

# Violin + boxplot with APR variants highlighted as red points
p <- ggplot(delta, aes(x = variant_group, y = deltaScore)) +
  geom_violin(data = subset(delta, variant_group != "Missense"),
              fill = "grey85", color = "grey40", width = 0.9) +
  geom_violin(data = subset(delta, variant_group == "Missense"),
              fill = "lightblue", color = "grey40", width = 0.9) +
  geom_boxplot(width = 0.15, outlier.shape = NA, fill = "white") +
  # Shaded null band (95% CI of synonymous + nonsense median)
  geom_rect(
    data = data.frame(xmin = 0.5, xmax = 2.5, ymin = null_lower, ymax = null_upper),
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    fill = "grey80", alpha = 0.2, inherit.aes = FALSE
  ) +
  geom_hline(yintercept = null_median, linetype = "dashed") +
  geom_hline(yintercept = 0,           linetype = "dotted") +
  # APR variants as filled red circles
  geom_point(
    data = subset(delta, variant_group == "Missense"),
    aes(fill = Impact == "APR"), shape = 21, color = "black", size = 2.5
  ) +
  scale_fill_manual(values = c("FALSE" = "white", "TRUE" = "red"), guide = "none") +
  labs(x = NULL, y = "Delta functional score (baseline – mercury)",
       title = "Delta score distributions with APR variants highlighted") +
  theme_classic() +
  theme(axis.text.x = element_text(size = 11), axis.title.y = element_text(size = 12))
print(p)

pdf("~/Desktop/APRdistribution.pdf", width = 8, height = 6); print(p); dev.off()

# Summary statistics
total_baseline_miss <- length(baseline_miss_tolerated)
num_APR             <- sum(delta$Impact == "APR", na.rm = TRUE)
percent_APR         <- 100 * num_APR / total_baseline_miss
total_missense      <- sum(delta$variant_type == "Missense")
num_negative        <- sum(delta$deltaScore < 0 & delta$variant_type == "Missense", na.rm = TRUE)

cat(sprintf("APR variants (baseline tolerated, significant negative delta): %d / %d (%.2f%%)\n",
            num_APR, total_baseline_miss, percent_APR))
cat(sprintf("Percentage of missense variants with negative deltaScore: %.2f%%\n",
            100 * num_negative / total_missense))

# Wilcoxon test: all missense delta scores vs. null (synonymous + nonsense)
missense_scores <- delta[variant_type == "Missense", deltaScore]
null_scores     <- delta[variant_group == "Synonymous + Nonsense", deltaScore]
wilcox_all_vs_null <- wilcox.test(missense_scores, null_scores, alternative = "two.sided")

cat(sprintf("Median deltaScore (Missense): %.3f\n", median(missense_scores, na.rm = TRUE)))
cat(sprintf("Median deltaScore (Null: Synonymous + Nonsense): %.3f\n", median(null_scores, na.rm = TRUE)))
print(wilcox_all_vs_null)
```

```{r, warning=FALSE, SNV Accessibility and Mercury Impact}
# =============================================================================
# CHUNK 14b — SNV Accessibility and Mercury Impact
#
# This analysis determines:
# 1. Which amino acid substitutions are SNV-reachable (single nucleotide 
#    variant accessible based on the genetic code)
# 2. Of those SNV-reachable substitutions, what fraction show statistically 
#    significant mercury effect
# Output: Percentage of SNV-accessible substitutions with mercury impact
# Data source: "Delta Map" sheet from Manuscript_Data.xlsx
# =============================================================================
# ─────────────────────────────────────────────────────────────────────────
# PART 1: Build SNV Reachability Matrix
# ─────────────────────────────────────────────────────────────────────────
# The genetic code: all codons and their amino acids

genetic_code <- list(
  "TTT" = "Phe", "TTC" = "Phe", "TTA" = "Leu", "TTG" = "Leu",
  "TCT" = "Ser", "TCC" = "Ser", "TCA" = "Ser", "TCG" = "Ser",
  "TAT" = "Tyr", "TAC" = "Tyr", "TAA" = "STOP", "TAG" = "STOP",
  "TGT" = "Cys", "TGC" = "Cys", "TGA" = "STOP", "TGG" = "Trp",
  "CTT" = "Leu", "CTC" = "Leu", "CTA" = "Leu", "CTG" = "Leu",
  "CCT" = "Pro", "CCC" = "Pro", "CCA" = "Pro", "CCG" = "Pro",
  "CAT" = "His", "CAC" = "His", "CAA" = "Gln", "CAG" = "Gln",
  "CGT" = "Arg", "CGC" = "Arg", "CGA" = "Arg", "CGG" = "Arg",
  "ATT" = "Ile", "ATC" = "Ile", "ATA" = "Ile", "ATG" = "Met",
  "ACT" = "Thr", "ACC" = "Thr", "ACA" = "Thr", "ACG" = "Thr",
  "AAT" = "Asn", "AAC" = "Asn", "AAA" = "Lys", "AAG" = "Lys",
  "AGT" = "Ser", "AGC" = "Ser", "AGA" = "Arg", "AGG" = "Arg",
  "GTT" = "Val", "GTC" = "Val", "GTA" = "Val", "GTG" = "Val",
  "GCT" = "Ala", "GCC" = "Ala", "GCA" = "Ala", "GCG" = "Ala",
  "GAT" = "Asp", "GAC" = "Asp", "GAA" = "Glu", "GAG" = "Glu",
  "GGT" = "Gly", "GGC" = "Gly", "GGA" = "Gly", "GGG" = "Gly"
)

# Function: Check if two codons differ by exactly 1 nucleotide
codons_differ_by_one <- function(codon1, codon2) {
  if (nchar(codon1) != 3 || nchar(codon2) != 3) return(FALSE)
  
  diffs <- 0
  for (pos in 1:3) {
    if (substr(codon1, pos, pos) != substr(codon2, pos, pos)) {
      diffs <- diffs + 1
    }
  }
  
  return(diffs == 1)
}

# Function: Check if an amino acid pair is SNV-reachable
# Input: from_aa, to_aa (3-letter codes, e.g., "Ala", "Arg")
# Output: TRUE if ANY codon pair differs by exactly 1 nucleotide
is_snv_reachable_pair <- function(from_aa, to_aa) {
  
  # Get all codons for FROM and TO amino acids
  from_codons <- names(genetic_code[which(genetic_code == from_aa)])
  to_codons <- names(genetic_code[which(genetic_code == to_aa)])
  
  # If either amino acid doesn't exist in genetic code, return FALSE
  if (length(from_codons) == 0 || length(to_codons) == 0) {
    return(FALSE)
  }
  
  # Test every FROM codon against every TO codon
  # If ANY pair differs by exactly 1 nucleotide, return TRUE
  for (from_codon in from_codons) {
    for (to_codon in to_codons) {
      if (codons_differ_by_one(from_codon, to_codon)) {
        return(TRUE)
      }
    }
  }
  
  # No SNV path found between any codon pairs
  return(FALSE)
}

# Pre-compute SNV reachability for all amino acid pairs
# This lookup table lets us quickly determine if any given substitution
# can be created with a single nucleotide change

cat("Building SNV reachability matrix...\n")

snv_reachable_pairs <- data.frame()

all_aa <- unique(unlist(genetic_code))
all_aa <- all_aa[all_aa != "STOP"]

# Test every FROM → TO pair (20 × 19 = 380 combinations)
for (from_aa in all_aa) {
  for (to_aa in all_aa) {
    if (from_aa != to_aa) {
      is_reachable <- is_snv_reachable_pair(from_aa, to_aa)
      snv_reachable_pairs <- rbind(snv_reachable_pairs, data.frame(
        from_aa = from_aa,
        to_aa = to_aa,
        is_snv_reachable = is_reachable,
        stringsAsFactors = FALSE
      ))
    }
  }
}

# Summary of reachability
reachable_count <- sum(snv_reachable_pairs$is_snv_reachable)
cat(sprintf("SNV Reachability Matrix Built:\n"))
cat(sprintf("  Total amino acid pairs tested: %d\n", nrow(snv_reachable_pairs)))
cat(sprintf("  SNV-reachable pairs: %d (%.1f%%)\n", reachable_count,
            100 * reachable_count / nrow(snv_reachable_pairs)))
cat("\n")

# ─────────────────────────────────────────────────────────────────────────
# PART 2: Load Delta Data from Excel
# ─────────────────────────────────────────────────────────────────────────
# Read the "Delta Map" sheet from Manuscript_Data.xlsx
# This sheet contains all variants with their environmental impact scores
# (significant column) and delta scores

delta_complete <- read_excel("~/Desktop/Manuscript_Data.xlsx", 
                              sheet = "Delta Map",
                              skip = 1) %>%  # Skip the description row
  as.data.table()
setDT(delta_complete)

# Ensure numeric columns are actually numeric
delta_complete[, `:=`(
  deltaScore = as.numeric(deltaScore),
  deltaSE = as.numeric(deltaSE),
  significant = as.logical(significant)
)]

cat(sprintf("Loaded Delta Map sheet: %d rows\n", nrow(delta_complete)))
cat("\n")

# ─────────────────────────────────────────────────────────────────────────
# PART 3: Parse HGVS Notation to Extract Amino Acids
# ─────────────────────────────────────────────────────────────────────────
# HGVS format: p.XxxnnYyy
#   Xxx = FROM amino acid (3-letter code)
#   nn = position number
#   Yyy = TO amino acid (3-letter code, or Ter for nonsense, or = for synonymous)
#
# Example: p.Ala110Arg means Alanine at position 110 → Arginine

delta_complete[, `:=`(
  from_aa = NA_character_,
  to_aa = NA_character_,
  position = NA_integer_,
  is_synonymous = FALSE,
  is_nonsense = FALSE
)]

# Function to parse HGVS and extract amino acid info
parse_hgvs_protein <- function(hgvs_str) {
  # Handle synonymous (p.Ala110=)
  if (grepl("=$", hgvs_str)) {
    from_aa <- regmatches(hgvs_str, regexpr("[A-Z][a-z]{2}", hgvs_str))
    pos <- regmatches(hgvs_str, regexpr("\\d+", hgvs_str))
    return(list(from = from_aa, to = from_aa, pos = as.numeric(pos), 
                is_syn = TRUE, is_stop = FALSE))
  }
  
  # Handle nonsense (p.Ala110Ter)
  if (grepl("Ter$", hgvs_str)) {
    from_aa <- regmatches(hgvs_str, regexpr("[A-Z][a-z]{2}(?=\\d)", hgvs_str, perl = TRUE))
    pos <- regmatches(hgvs_str, regexpr("\\d+(?=[A-Z])", hgvs_str, perl = TRUE))
    return(list(from = from_aa, to = "Ter", pos = as.numeric(pos),
                is_syn = FALSE, is_stop = TRUE))
  }
  
  # Handle missense (p.Ala110Arg)
  from_aa <- regmatches(hgvs_str, regexpr("[A-Z][a-z]{2}(?=\\d)", hgvs_str, perl = TRUE))
  pos <- regmatches(hgvs_str, regexpr("\\d+(?=[A-Z])", hgvs_str, perl = TRUE))
  to_aa <- regmatches(hgvs_str, regexpr("[A-Z][a-z]{2}$", hgvs_str, perl = TRUE))
  
  return(list(from = from_aa, to = to_aa, pos = as.numeric(pos),
              is_syn = FALSE, is_stop = FALSE))
}

# Apply parsing to all variants
for (i in 1:nrow(delta_complete)) {
  parsed <- parse_hgvs_protein(delta_complete[i, hgvs_pro])
  delta_complete[i, `:=`(
    from_aa = parsed$from,
    to_aa = parsed$to,
    position = parsed$pos,
    is_synonymous = parsed$is_syn,
    is_nonsense = parsed$is_stop
  )]
}

cat(sprintf("Parsed HGVS notation for all %d variants\n", nrow(delta_complete)))
cat("\n")

# ─────────────────────────────────────────────────────────────────────────
# PART 4: Classify SNV Reachability
# ─────────────────────────────────────────────────────────────────────────
# For each missense variant:
#   1. Extract FROM and TO amino acids
#   2. Look up in SNV reachability table
#   3. Mark as SNV-reachable if ANY codon pair could differ by 1 SNV

delta_complete[, is_snv_reachable := FALSE]

for (i in 1:nrow(delta_complete)) {
  from_aa <- delta_complete[i, from_aa]
  to_aa <- delta_complete[i, to_aa]
  
  # Only classify missense variants (not synonymous or nonsense)
  if (!is.na(from_aa) && !is.na(to_aa) && 
      from_aa != to_aa && 
      to_aa != "Ter" &&
      !delta_complete[i, is_synonymous] &&
      !delta_complete[i, is_nonsense]) {
    
    # Look up reachability in pre-computed table
    match <- snv_reachable_pairs[snv_reachable_pairs$from_aa == from_aa &
                                   snv_reachable_pairs$to_aa == to_aa, ]
    
    if (nrow(match) > 0 && match$is_snv_reachable[1]) {
      delta_complete[i, is_snv_reachable := TRUE]
    }
  }
}

cat("SNV reachability classified for all variants\n")
cat("\n")

# ─────────────────────────────────────────────────────────────────────────
# PART 5: Calculate Environmental Impact Statistics
# ─────────────────────────────────────────────────────────────────────────

# Count total SNV-reachable missense variants
total_snv_reachable <- sum(delta_complete[, is_snv_reachable], na.rm = TRUE)

# Count SNV-reachable variants with SIGNIFICANT environmental impact
# (significant == TRUE means mercury effect is statistically significant)
snv_with_significant_impact <- sum(
  delta_complete[is_snv_reachable == TRUE & significant == TRUE, .N]
)

# Calculate percentage
pct_snv_with_impact <- if (total_snv_reachable > 0) {
  100 * snv_with_significant_impact / total_snv_reachable
} else {
  NA_real_
}

# ─────────────────────────────────────────────────────────────────────────
# PART 6: Print Summary Statistics
# ─────────────────────────────────────────────────────────────────────────

cat("=================================================================\n")
cat("SNV ACCESSIBILITY AND ENVIRONMENTAL IMPACT ANALYSIS\n")
cat("=================================================================\n\n")

# Overall variant counts
total_variants <- nrow(delta_complete)
missense_count <- sum(!delta_complete[, is_synonymous] & 
                      !delta_complete[, is_nonsense], na.rm = TRUE)
synonymous_count <- sum(delta_complete[, is_synonymous], na.rm = TRUE)
nonsense_count <- sum(delta_complete[, is_nonsense], na.rm = TRUE)

cat("VARIANT TYPE DISTRIBUTION:\n")
cat(sprintf("  Total variants in delta data: %d\n", total_variants))
cat(sprintf("  Missense variants: %d (%.1f%%)\n", missense_count, 
            100 * missense_count / total_variants))
cat(sprintf("  Synonymous variants: %d (%.1f%%)\n", synonymous_count, 
            100 * synonymous_count / total_variants))
cat(sprintf("  Nonsense variants: %d (%.1f%%)\n", nonsense_count, 
            100 * nonsense_count / total_variants))
cat("\n")

# SNV reachability statistics
cat("SNV REACHABILITY (among missense variants):\n")
cat(sprintf("  SNV-reachable substitutions: %d / %d (%.1f%%)\n",
            total_snv_reachable, missense_count,
            100 * total_snv_reachable / missense_count))
cat(sprintf("  Non-SNV-reachable substitutions: %d / %d (%.1f%%)\n",
            missense_count - total_snv_reachable, missense_count,
            100 * (missense_count - total_snv_reachable) / missense_count))
cat("\n")

# Environmental impact among SNV-reachable
cat("KEY RESULT: ENVIRONMENTAL IMPACT (MERCURY EFFECT)\n")
cat("Among SNV-reachable substitutions:\n")
cat(sprintf("  With significant environmental impact: %d / %d (%.2f%%)\n",
            snv_with_significant_impact, total_snv_reachable, pct_snv_with_impact))
cat(sprintf("  Without significant impact: %d / %d (%.2f%%)\n",
            total_snv_reachable - snv_with_significant_impact, total_snv_reachable,
            100 - pct_snv_with_impact))
cat("\n")

# Breakdown by delta score direction
snv_negative_delta <- sum(
  delta_complete[is_snv_reachable == TRUE & significant == TRUE & deltaScore < 0, .N],
  na.rm = TRUE
)
snv_positive_delta <- sum(
  delta_complete[is_snv_reachable == TRUE & significant == TRUE & deltaScore > 0, .N],
  na.rm = TRUE
)

if (snv_with_significant_impact > 0) {
  cat("DIRECTIONALITY (among SNV-reachable with impact):\n")
  cat(sprintf("  Negative delta (reduced function in Hg): %d (%.1f%%)\n",
              snv_negative_delta, 100 * snv_negative_delta / snv_with_significant_impact))
  cat(sprintf("  Positive delta (improved function in Hg): %d (%.1f%%)\n",
              snv_positive_delta, 100 * snv_positive_delta / snv_with_significant_impact))
  cat("\n")
}

# Summary box
cat("=================================================================\n")
cat(sprintf("SUMMARY:\n"))
cat(sprintf("  %.2f%% of SNV-accessible amino acid substitutions\n", pct_snv_with_impact))
cat(sprintf("  show statistically significant environmental impact (mercury)\n"))
cat("=================================================================\n\n")

# Export detailed results
results_export <- delta_complete[is_snv_reachable == TRUE, 
                                 .(hgvs_pro, from_aa, to_aa, position,
                                   deltaScore, deltaSE, significant)]
results_export <- results_export[order(position, from_aa, to_aa)]

write.csv(results_export, "~/Desktop/CPOX_SNV_Environmental_Impact.csv", row.names = FALSE)
cat(sprintf("Detailed results exported to: CPOX_SNV_Environmental_Impact.csv\n"))
cat(sprintf("  Total SNV-reachable rows: %d\n", nrow(results_export)))
```

```{r, warning=FALSE, Delta Significance Heatmap}
# =============================================================================
# CHUNK 15 — Delta Significance Heatmap (Variant-Level)
#
# Produces a full amino acid × position tile heatmap coloured by variant impact
# category (APR / Improved / Known / neutral / Synonymous), with a stacked
# consensus track above showing the fraction of each category per position.
#
# Y-axis order follows the Kyte-Doolittle hydrophobicity scale (AVLIMFYWRHKDESTNQGCP).
# =============================================================================

delta <- read_sheet("Delta Map")
setDT(delta)
delta <- delta[is.finite(deltaScore) & is.finite(deltaSE)]

delta[, variant_type  := "Missense"]
delta[grepl("Ter$", hgvs_pro), variant_type := "Nonsense"]
delta[grepl("=$",   hgvs_pro), variant_type := "Synonymous"]
delta[, variant_group := variant_type]
delta[variant_type %in% c("Synonymous","Nonsense"), variant_group := "Synonymous + Nonsense"]

baseline_map <- read_sheet("Baseline Map")
setDT(baseline_map)
setnames(baseline_map, 1:4, c("hgvs_pro","baseline_score","se","df"))
baseline_map[, baseline_lower := baseline_score + se]
baseline_map[, baseline_stop  := grepl("Ter$", hgvs_pro)]
baseline_map[, baseline_syn   := grepl("=$",   hgvs_pro)]
baseline_map[, baseline_miss  := !(baseline_stop | baseline_syn)]
baseline_miss_tolerated <- baseline_map[baseline_miss & baseline_lower >= 0.5, hgvs_pro]

delta[, Impact := "neutral"]
APR_set      <- delta[significant == TRUE  & deltaScore < 0 & hgvs_pro %in% baseline_miss_tolerated, hgvs_pro]
known_set    <- delta[significant == FALSE & hgvs_pro %in% baseline_miss_tolerated, hgvs_pro]
improved_set <- delta[significant == TRUE  & deltaScore > 0 & hgvs_pro %in% baseline_miss_tolerated, hgvs_pro]
delta[hgvs_pro %in% known_set,    Impact := "known"]
delta[hgvs_pro %in% APR_set,      Impact := "APR"]
delta[hgvs_pro %in% improved_set, Impact := "improved"]

# Single-letter amino acid lookup (for y-axis labels)
aa3to1 <- c(
  Ala="A",Arg="R",Asn="N",Asp="D",Cys="C",Gln="Q",Glu="E",Gly="G",
  His="H",Ile="I",Leu="L",Lys="K",Met="M",Phe="F",Pro="P",Ser="S",
  Thr="T",Trp="W",Tyr="Y",Val="V"
)

# Build per-variant heatmap data.frame
heatmap_df <- delta %>%
  mutate(
    pos             = as.numeric(str_extract(hgvs_pro, "[0-9]+")),
    substitution_AA = case_when(
      grepl("=$", hgvs_pro) ~ str_sub(hgvs_pro, 3, 3),          # synonymous: keep original AA
      TRUE                  ~ aa3to1[str_sub(hgvs_pro, -3, -1)]  # missense: extract destination AA
    )
  ) %>%
  mutate(
    color_bin = case_when(
      grepl("=$", hgvs_pro) ~ "Synonymous",
      Impact == "known"     ~ "known",
      Impact == "APR"       ~ "APR",
      Impact == "neutral"   ~ "neutral",
      Impact == "improved"  ~ "improved",
      TRUE                  ~ NA_character_
    )
  )

# Y-axis order (Kyte-Doolittle hydrophobicity)
aa_order <- str_split("AVLIMFYWRHKDESTNQGCP", "", simplify = TRUE)

plot_df <- heatmap_df %>%
  filter(!is.na(pos) & !is.na(substitution_AA)) %>%
  mutate(
    substitution_AA = factor(substitution_AA, levels = rev(aa_order)),
    color_bin       = factor(color_bin, levels = c("known","APR","neutral","improved","Synonymous"))
  )

# Build stacked consensus track (fractional composition per position)
consensus_df <- heatmap_df %>%
  group_by(pos, color_bin) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(pos) %>%
  mutate(
    frac = count / sum(count),
    ymin = 22 + cumsum(lag(frac, default = 0)),
    ymax = 22 + cumsum(frac)
  ) %>%
  ungroup()

impact_colors <- c(
  "known"      = "#AEBBE6",
  "APR"        = "#CDEED6",
  "neutral"    = "#F7F8FD",
  "improved"   = "#DFA6A5",
  "Synonymous" = "#FCED98"
)

heatmap_plot <- ggplot() +
  # Main heatmap tiles
  geom_tile(data = plot_df, aes(x = pos, y = substitution_AA, fill = color_bin),
            width = 1, height = 1) +
  scale_fill_manual(values = impact_colors, na.value = "grey80") +
  # Stacked consensus track above the heatmap (y positions 22+)
  geom_rect(data = consensus_df,
            aes(xmin = pos - 0.5, xmax = pos + 0.5, ymin = ymin, ymax = ymax, fill = color_bin),
            inherit.aes = FALSE) +
  labs(x = "Amino acid position", y = "Substituted residue",
       title = "CPOX Variant Heatmap with Stacked Consensus Track") +
  scale_x_continuous(breaks = seq(1, max(plot_df$pos), by = 5)) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text  = element_text(size = 4),
        axis.title = element_text(size = 12),
        plot.title = element_text(hjust = 0.5))

pdf("~/Desktop/CPOX_heatmap_stacked_consensus.pdf", width = 14, height = 3)
print(heatmap_plot); dev.off()
```


```{r, warning=FALSE, Balanced Precision vs Recall Curves for Maps}
# =============================================================================
# CHUNK 16 — Balanced Precision-Recall Curves (PRC)
#
# Evaluates the clinical discriminative performance of:
#   • CPOX baseline DMS map
#   • CPOX metal DMS map
#   • AlphaMissense
#   • ESM-1v
#   • VARITY_R
#
# Positive class: HCP / HP variants (hereditary coproporphyria / harderoporphyria)
# Negative class: disease_type == "none" (benign)
# VUS and conflicting variants are excluded.
#
# "Balanced" PRC accounts for class imbalance.
# DMS scores are negated before PRC (lower score = more pathogenic in DMS convention).
#
# Four PRC subsets:
#   1. All variants
#   2. Excluding HP (harderoporphyria — atypical severe presentation)
#   3. Excluding MTS positions (< 110) — these may behave differently
#   4. Excluding HCP
#
# A boxplot of scores by clinical annotation (B/LB vs P/LP) is also produced,
# with Wilcoxon test between groups for each map.
# =============================================================================

generate_prc_plots <- function(ref_file, baseline_file, metal_file = NULL) {
  library(data.table); library(dplyr); library(ggplot2); library(tidyr)
  
  # Load and label reference data
  ref_data <- read_sheet("ReferenceSet") %>%
    mutate(
      Pathogenic = case_when(
        disease_type %in% c("HCP","HP") ~ "pathogenic",
        disease_type == "none"          ~ "benign",
        TRUE                            ~ disease_type
      )
    )
  
  baseline_data <- read_sheet("Baseline Map")
  colnames(baseline_data)[2] <- "baseline"
  
  if (!is.null(metal_file)) {
    metal_data <- read_sheet("Mercury Map")
    colnames(metal_data)[2] <- "metal"
  }
  
  complete_dataset <- merge(ref_data, baseline_data, by = "hgvs_pro")
  if (!is.null(metal_file)) {
    complete_dataset <- merge(complete_dataset, metal_data, by = "hgvs_pro")
  } else {
    complete_dataset$metal <- NA
  }
  
  complete_dataset <- complete_dataset %>%
    select(hgvs_pro, baseline, AlphaMissense, ESM.1v, VARITY_R, disease_type, metal) %>%
    mutate(pos = as.numeric(gsub("[^0-9]", "", hgvs_pro)))
  
  # Clinical annotation boxplot
  variant_distribution <- complete_dataset %>%
    mutate(
      color       = ifelse(disease_type == "HP", "red", "grey30"),
      disease_type = case_when(
        disease_type %in% c("HCP","HP") ~ "P/LP",
        disease_type == "none"          ~ "B/LB",
        TRUE                            ~ disease_type
      )
    )
  
  variant_long <- variant_distribution %>%
    pivot_longer(cols = c(baseline, metal), names_to = "ScoreType", values_to = "Score")
  
  annot_plot <- ggplot(variant_long, aes(x = disease_type, y = Score, fill = ScoreType)) +
    geom_boxplot(alpha = 0.3, position = position_dodge(width = 0.8), outlier.shape = NA) +
    geom_jitter(aes(color = color, group = ScoreType), size = 2, alpha = 0.8,
                position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.8)) +
    scale_fill_manual(values = c(baseline = "#11EE89", metal = "#EE1176")) +
    scale_color_identity() +
    theme_minimal() +
    xlab("Clinical Annotation") + ylab("Score") +
    labs(title = "Baseline vs Metal Scores by Clinical Annotation", fill = "Map") +
    theme(legend.position = "right", axis.text = element_text(size = 12),
          axis.title = element_text(size = 14))
  print(annot_plot)
  
  # Wilcoxon test: P/LP vs. B/LB for each map
  for (score_type in c("baseline","metal")) {
    group_B_LB <- variant_distribution %>% filter(disease_type == "B/LB") %>% pull(all_of(score_type))
    group_P_LP <- variant_distribution %>% filter(disease_type == "P/LP") %>% pull(all_of(score_type))
    test_result <- wilcox.test(group_P_LP, group_B_LB, exact = FALSE)
    median_diff <- median(group_P_LP, na.rm = TRUE) - median(group_B_LB, na.rm = TRUE)
    cat(sprintf("Score type: %s\nMedian difference (P/LP - B/LB): %.6f\n", score_type, median_diff))
    if (test_result$p.value < 2.2e-16) {
      cat("P-value: < 2.2e-16\n\n")
    } else {
      cat(sprintf("P-value: %.8f\n\n", test_result$p.value))
    }
  }
  
  # Prepare PRC dataset: negate scores (lower DMS score = more pathogenic)
  PRC_data <- complete_dataset %>%
    filter(
      !grepl("VUS",        disease_type, ignore.case = TRUE) &
      !grepl("conflicting",disease_type, ignore.case = TRUE)
    ) %>%
    mutate(
      Pathogenic = case_when(
        disease_type %in% c("HCP","HP") ~ TRUE,
        disease_type == "none"          ~ FALSE,
        TRUE                            ~ FALSE
      ),
      baseline = -baseline,
      metal    = ifelse(!is.na(metal), -metal, NA)
    )
  print(PRC_data)
  
  # PRC plotting helper using yogiroc
  plot_prc <- function(data, title) {
    prc_data <- data %>%
      select(Pathogenic, baseline, AlphaMissense, ESM.1v, VARITY_R, metal) %>%
      mutate(Pathogenic = as.logical(Pathogenic))
    
    if (all(is.na(prc_data$metal))) {
      prc_data <- prc_data %>% select(-metal)
      colors   <- c("#11EE89","#D02F43","#2FD0BC","black")
    } else {
      colors   <- c("#11EE89","#D02F43","#2FD0BC","black","#EE1176")
    }
    
    yrobj <- yr2(truth = prc_data$Pathogenic, scores = prc_data[, -1])
    draw.prc(yrobj, balanced = TRUE, main = title, col = colors)
    cat("Total Pathogenic (TRUE) and Benign (FALSE):\n")
    print(table(prc_data$Pathogenic))
  }
  
  # Generate all PRC subsets
  plot_prc(PRC_data, "All Variants")
  plot_prc(PRC_data %>% filter(!grepl("HP",  disease_type, ignore.case = TRUE)), "Excluding HP Variants")
  plot_prc(PRC_data %>% filter(as.numeric(gsub("[^0-9]","",hgvs_pro)) >= 110),   "Excluding MTS Positions (2-110)")
  plot_prc(PRC_data %>% filter(!grepl("HCP", disease_type, ignore.case = TRUE)), "Excluding HCP Variants")
  
  # Save PDFs
  pdf("~/Desktop/All_PRC.pdf", width = 8, height = 6)
  plot_prc(PRC_data, "All Variants")
  ggsave("~/Desktop/All_ClinicalVariants.pdf", plot = annot_plot, width = 8, height = 6)
  dev.off()
  pdf("~/Desktop/PRC_noHP.pdf",  8, 6)
  plot_prc(PRC_data %>% filter(!grepl("HP",  disease_type, ignore.case = TRUE)), "Excluding HP Variants"); dev.off()
  pdf("~/Desktop/PRC_noHCP.pdf", 8, 6)
  plot_prc(PRC_data %>% filter(!grepl("HCP", disease_type, ignore.case = TRUE)), "Excluding HCP Variants"); dev.off()
  pdf("~/Desktop/PRC_noMTS.pdf", 8, 6)
  plot_prc(PRC_data %>% filter(as.numeric(gsub("[^0-9]","",hgvs_pro)) >= 110),   "Excluding MTS Positions (2-110)"); dev.off()
}

generate_prc_plots(
  ref_file      = "~/Desktop/Manuscript_Data.xlsx",
  baseline_file = "~/Desktop/Manuscript_Data.xlsx",
  metal_file    = "~/Desktop/Manuscript_Data.xlsx"
)
```


```{r, Balanced Precision vs Recall Curve for Random Predictor, warning=FALSE}
# =============================================================================
# CHUNK 17 — PRC with Random Predictor
#
# Adds a random predictor (shuffled truth labels) to the PRC to provide a
# chance-level baseline. This confirms that the observed PRC area under the
# curve exceeds random performance.
# seed = 123 ensures reproducibility of the random shuffle.
# =============================================================================

library(data.table); library(dplyr); library(ggplot2); library(tidyr); library(yogiroc)

generate_prc_plots <- function(ref_file, baseline_file, metal_file = NULL) {
  
  ref_data <- read_sheet("ReferenceSet")
  
  baseline_data <- read_sheet("Baseline Map")
  colnames(baseline_data)[2] <- "baseline"
  
  if (!is.null(metal_file)) {
    metal_data <- read_sheet("Mercury Map")
    colnames(metal_data)[2] <- "metal"
  }
  
  complete_dataset <- merge(ref_data, baseline_data, by = "hgvs_pro")
  if (!is.null(metal_file)) {
    complete_dataset <- merge(complete_dataset, metal_data, by = "hgvs_pro")
  } else {
    complete_dataset$metal <- NA
  }
  
  # Prepare PRC data (negate scores, filter VUS/conflicting)
  PRC_data <- complete_dataset %>%
    filter(
      !grepl("VUS",         disease_type, ignore.case = TRUE) &
      !grepl("conflicting", disease_type, ignore.case = TRUE)
    ) %>%
    mutate(
      Pathogenic = case_when(
        disease_type %in% c("HCP","HP") ~ TRUE,
        disease_type == "none"          ~ FALSE,
        TRUE                            ~ FALSE
      ),
      baseline = -baseline,
      metal    = ifelse(!is.na(metal), -metal, NA)
    ) %>%
    select(hgvs_pro, Pathogenic, baseline, AlphaMissense, ESM.1v, VARITY_R, metal)
  
  # Add random predictor: shuffled pathogenicity labels (set.seed for reproducibility)
  set.seed(123)
  PRC_data$Random <- -1 * as.numeric(sample(PRC_data$Pathogenic))
  
  cat("\n===== PRC DATA TABLE USED =====\n")
  print(PRC_data)
  cat("\nPathogenic counts:\n")
  print(table(PRC_data$Pathogenic))
  
  score_matrix <- PRC_data %>%
    select(baseline, AlphaMissense, ESM.1v, VARITY_R, metal, Random)
  
  colors <- c(
    "#11EE89",  # baseline
    "#D02F43",  # AlphaMissense
    "#2FD0BC",  # ESM.1v
    "black",    # VARITY_R
    "#EE1176",  # metal
    "grey50"    # Random
  )
  
  yrobj <- yr2(truth = PRC_data$Pathogenic, scores = score_matrix)
  
  pdf("~/Desktop/PRC_Random.pdf", width = 8, height = 6)
  draw.prc(yrobj, balanced = TRUE, main = "All Variants (Random Included)", col = colors)
  dev.off()
  
  cat("\nPRC figure saved to ~/Desktop/PRC_Random.pdf\n")
}

generate_prc_plots(
  ref_file      = "~/Desktop/Manuscript_Data.xlsx",
  baseline_file = "~/Desktop/Manuscript_Data.xlsx",
  metal_file    = "~/Desktop/Manuscript_Data.xlsx"
)
```


```{r, warning=FALSE, LLRp}
# =============================================================================
# CHUNK 18 — LLRp Evidence for VUS Reclassification
#
# Applies the Tavtigian et al. (2020) log-likelihood ratio (LLR) framework
# to quantify functional evidence strength for CPOX ClinVar VUS variants.
#
# LLR thresholds (ACMG/AMP evidence levels):
#   > 1.2723   : pathogenic very strong
#   0.636–1.272: pathogenic moderate
#   0.318–0.636: pathogenic supporting
#  -0.318–0.318: no evidence
#  -1.272–-0.318: benign supporting
#   < -1.272   : benign strong
#
# A stacked bar chart shows the evidence distribution per ClinVar category
# (P/LP, B/BP, VUS). Percentage of VUS with actionable evidence is reported.
# =============================================================================

LLRp_data <- read_sheet("LLRp Scores")

# Assign ACMG evidence categories based on LLR thresholds
LLRp_data$category <- dplyr::case_when(
  LLRp_data$llr >  0.636167   & LLRp_data$llr <  1.2723233 ~ "patho_moderate",
  LLRp_data$llr >  0.3180808  & LLRp_data$llr <  0.636167  ~ "patho_support",
  LLRp_data$llr <  0.3180808  & LLRp_data$llr > -0.3180808 ~ "none",
  LLRp_data$llr > -1.2723233  & LLRp_data$llr < -0.3180808 ~ "benign_support",
  LLRp_data$llr < -1.2723233                                ~ "benign_strong",
  TRUE                                                       ~ "patho_vstrong"
)

# Load and classify reference variants
ref_Set <- read_sheet("ReferenceSet")
ref_Set <- ref_Set %>%
  mutate(
    clinvar = case_when(
      grepl("HP|HCP", disease_type, ignore.case = TRUE) ~ "P/LP",
      grepl("none",   disease_type, ignore.case = TRUE) ~ "B/BP",
      TRUE                                              ~ "VUS"
    )
  )

new_evidence_data <- merge(LLRp_data, ref_Set, by = "hgvs_pro")
new_evidence_data <- new_evidence_data[, c("hgvs_pro","clinvar","llr","category")]

# Colour scheme: green (benign) → grey (none) → purple (pathogenic)
category_colors <- c(
  "benign_strong"  = "#50DD22",
  "benign_support" = "#A0FF80",
  "none"           = "grey50",
  "patho_support"  = "#D18BEE",
  "patho_moderate" = "#AF22DD",
  "patho_strong"   = "#9B00C8",
  "patho_vstrong"  = "#8000A0"
)

LLR_evidence <- ggplot(new_evidence_data, aes(x = clinvar, fill = category)) +
  geom_bar(position = position_stack(reverse = TRUE)) +
  scale_fill_manual(values = category_colors) +
  theme_minimal() +
  labs(x = "ClinVar Category", y = "Count", fill = "LLR Category") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(filename = "~/Desktop/LLR_evidence.pdf", plot = LLR_evidence, width = 8, height = 6)

# Percentage of VUS with any functional evidence (i.e. category ≠ "none")
vus_data          <- new_evidence_data[new_evidence_data$clinvar == "VUS", ]
total_vus         <- nrow(vus_data)
vus_with_none     <- nrow(vus_data[vus_data$category == "none", ])
vus_without_none  <- ((total_vus - vus_with_none) / total_vus) * 100

cat("Percentage of VUS entries without 'none' in category:", vus_without_none, "%\n")
```


```{r, warning=FALSE, VUS Numbers}
# =============================================================================
# CHUNK 19 — VUS Reclassification Summary
#
# Breaks down the functional evidence categories for the VUS subset.
# Reports counts and percentages (calculated excluding "none" from denominator
# to show the distribution among those with actual evidence).
# =============================================================================

# Counts per category for VUS variants
category_counts <- table(vus_data$category)

# Exclude "none" for percentage calculation (report among actionable evidence only)
categories_for_percentage <- category_counts[names(category_counts) != "none"]

category_percentages <- (categories_for_percentage / sum(categories_for_percentage)) * 100

category_summary <- data.frame(
  Category   = names(categories_for_percentage),
  Count      = as.vector(categories_for_percentage),
  Percentage = as.vector(category_percentages)
)

cat("Counts and Percentages of VUS entries by category (excluding 'none' from percentage):\n")
print(category_summary)
```


```{r, warning=FALSE, Variant Validation}
# =============================================================================
# CHUNK 20 — Variant Validation (Growth Curves)
#
# Analyses yeast growth curves from a 96-well plate reader experiment used
# to orthogonally validate a subset of DMS variants.
#
# Data source: 'Growth Curves Summary' sheet in Manuscript_Data.xlsx
#   Dataset='WT_N272'    : WT vs p.Asn272His across five mercury concentrations
#   Dataset='0 vs 96 µM' : WT, p.Asn272His, p.Leu155Trp, p.Gly188Gln at 0 and 96 µM
#
# Condition and Comparison columns are mercury concentrations in µM.
#
# Output: mean ± SD growth curves (linear and log scale) with ribbon shading.
# =============================================================================

DATA_PATH     <- "~/Desktop/Manuscript_Data.xlsx"
output_prefix <- "~/Desktop/CPOX_Growth_Curves"

# ── Read data ─────────────────────────────────────────────────────────────
gc <- read_excel(DATA_PATH, sheet = "Growth Curves Summary", skip = 1) %>%
  mutate(Time_hours = as.numeric(Time_hours))

df_summary <- gc %>%
  filter(Dataset == "WT_N272") %>%
  mutate(
    Variant    = factor(Variant, levels = c("WT", "N272")),
    Condition  = factor(Condition),
    Comparison = as.numeric(Comparison)
  )

df_AD <- gc %>%
  filter(Dataset == "0 vs 96 µM") %>%
  mutate(
    Variant   = factor(Variant, levels = unique(Variant)),
    Condition = factor(Condition)
  )

# ── Loop over comparison concentrations (WT_N272 dataset) ─────────────────
comparison_rows  <- unique(df_summary$Comparison)
plot_list_linear <- list()
plot_list_log    <- list()

for (comparison_conc in comparison_rows) {

  df_plot <- df_summary %>% filter(Comparison == comparison_conc)

  linetype_values <- setNames(c("dotted", "solid"), unique(as.character(df_plot$Condition)))
  color_values    <- RColorBrewer::brewer.pal(4, "Set1")[1:2]
  plot_key        <- paste0(comparison_conc, "uM")

  p_linear <- ggplot(df_plot, aes(Time_hours, mean_OD, color = Variant, linetype = Condition)) +
    geom_line(size = 1, aes(group = interaction(Variant, Condition))) +
    geom_ribbon(aes(ymin = mean_OD - sd_OD, ymax = mean_OD + sd_OD, fill = Variant),
                alpha = 0.15, color = NA) +
    scale_linetype_manual(values = linetype_values) +
    scale_color_manual(values = color_values) +
    scale_fill_manual(values = color_values) +
    labs(title = paste0("Growth Curves (Linear) — 0 vs ", comparison_conc, " µM"),
         x = "Time (hours)", y = "OD600") +
    theme_minimal()

  p_log <- ggplot(df_plot, aes(Time_hours, mean_OD, color = Variant, linetype = Condition)) +
    geom_line(size = 1, aes(group = interaction(Variant, Condition))) +
    geom_ribbon(aes(ymin = mean_OD - sd_OD, ymax = mean_OD + sd_OD, fill = Variant),
                alpha = 0.15, color = NA) +
    scale_linetype_manual(values = linetype_values) +
    scale_y_log10() +
    scale_color_manual(values = color_values) +
    scale_fill_manual(values = color_values) +
    labs(title = paste0("Growth Curves (Log scale) — 0 vs ", comparison_conc, " µM"),
         x = "Time (hours)", y = "OD600 (log scale)") +
    theme_minimal()

  ggsave(paste0(output_prefix, "_", plot_key, "_Linear.pdf"), p_linear, width = 8, height = 6)
  ggsave(paste0(output_prefix, "_", plot_key, "_Log.pdf"),    p_log,    width = 8, height = 6)

  plot_list_linear[[plot_key]] <- p_linear
  plot_list_log[[plot_key]]    <- p_log
}

# ── 0 vs 96 µM plot (all variants) ────────────────────────────────────────
linetype_values_AD <- setNames(c("dotted", "solid"), levels(df_AD$Condition))
color_values_AD    <- RColorBrewer::brewer.pal(4, "Set1")[1:length(unique(df_AD$Variant))]

p_AD_linear <- ggplot(df_AD, aes(Time_hours, mean_OD, color = Variant, linetype = Condition)) +
  geom_line(size = 1, aes(group = interaction(Variant, Condition))) +
  geom_ribbon(aes(ymin = mean_OD - sd_OD, ymax = mean_OD + sd_OD, fill = Variant),
              alpha = 0.15, color = NA) +
  scale_linetype_manual(values = linetype_values_AD) +
  scale_color_manual(values = color_values_AD) +
  scale_fill_manual(values = color_values_AD) +
  labs(title = "Growth Curves (Linear) — 0 vs 96 µM (All Variants)",
       x = "Time (hours)", y = "OD600") +
  theme_minimal()

p_AD_log <- ggplot(df_AD, aes(Time_hours, mean_OD, color = Variant, linetype = Condition)) +
  geom_line(size = 1, aes(group = interaction(Variant, Condition))) +
  geom_ribbon(aes(ymin = mean_OD - sd_OD, ymax = mean_OD + sd_OD, fill = Variant),
              alpha = 0.15, color = NA) +
  scale_linetype_manual(values = linetype_values_AD) +
  scale_y_log10() +
  scale_color_manual(values = color_values_AD) +
  scale_fill_manual(values = color_values_AD) +
  labs(title = "Growth Curves (Log scale) — 0 vs 96 µM (All Variants)",
       x = "Time (hours)", y = "OD600 (log scale)") +
  theme_minimal()

plot_list_linear[["0_vs_96uM_all"]] <- p_AD_linear
plot_list_log[["0_vs_96uM_all"]]    <- p_AD_log

# ── Combined grid ─────────────────────────────────────────────────────────
combined_linear <- wrap_plots(plot_list_linear, ncol = 3)
combined_log    <- wrap_plots(plot_list_log,    ncol = 3)

ggsave(paste0(output_prefix, "_Combined_Linear.pdf"), combined_linear, width = 16, height = 8)
ggsave(paste0(output_prefix, "_Combined_Log.pdf"),    combined_log,    width = 16, height = 8)
```
